---
title: "BAMSE DBS"
author: "AB"
format: html
editor: visual
---

## Load packages and functions:

```{r}
library(tidyverse)
library(OlinkAnalyze)
library(ggbeeswarm)
library(ggpubr) 
library(rstatix)i
library(patchwork)
library(emmeans)
library(readxl)
library(ghibli)
library(lubridate)


## Load funcitons:
source("plot_dimred.R")
source("comp_groups.R")

```

## Load data (Luminex, Olink and sample information):

```{r}
serology_norm <- read.csv("BAMSE_serology_log2_bb_igm_normalized_data.csv")

samp_info_key <- read.csv("./datafiles/sample_information/bamse_sampleid_key_v2.csv")
sample_info_key <- samp_info_key

cluster_5prot_info <- ("cluster_information_bamse_subjects_hierarchical5protein_maximumdist.csv")

empty_disc_list <- read_excel("empty_disc_list.xlsx")

olink_bamse_protpqn_blank_rm_intnorm <- read.csv("protPQN intensity normalized bamse data_241004.csv")


## Sample information:
sample_info <- read.csv("normalized serology data with sample information_bamse samples_240627.csv")
sample_info <- sample_info[,c(2:38,110:122) ]

# Wide sample info:
sample_info_wide <- read.csv("BAMSE C19 2023-03-06_v2.csv")

## Updated sampling dates:
sampling_date_updated <- read.csv("./sample_data/Datum_Filterpapper_korrigerad_250227.csv")

### Cell count data:
bc_data <- read_excel("BAMSE_C19_Blodstatus.xlsx")
bc_data <- bc_data %>%
  mutate(across(where(is.character), ~ str_replace_all(., "(?<!\\d)\\.(\\d)", "0.\\1")))

bc_data <- bc_data %>%
  mutate(across(c(basgran_c19, eosgran_c19, erytrocyt_c19, evf_c19, leukocyt_c19, lymfocyt_c19,  monocyt_c19, neutgran_c19 ), as.numeric))


## Remove the samples with empty discs from the serology data:
empty_disc_samp <- empty_disc_list |>
  filter(No_blood == 1) |>
  pull(sample_name...8). # Containt the sample names

serology_norm_rm_empty <- serology_norm |>
  rename(sample_name = X) |>
  filter(!sample_name %in% empty_disc_samp)

## Combine data with sample information:
serology_norm_rm_empty_sinfo <- serology_norm_rm_empty |>
  left_join(sample_info, by = "sample_name") |>
  mutate(sampling_phase = plot_variable) # Change name of plot_variable to sampling_phase to make it more clear

serology_norm_rm_empty_sinfo <- serology_norm_rm_empty_sinfo %>%
  left_join(sample_info_wide[ , c("idnr", "sminet_date", "IgG_bamse")], by = "idnr") |>
  mutate(
    sample_date = as.Date(sample_date),
    sminet_date = as.Date(sminet_date)
  ) |>
  mutate(time_diff = ifelse(sminet_date < sample_date, sample_date - sminet_date, NA))

beads_of_interest <- colnames(serology_norm_rm_empty)[c(2:73)]
covid_beads <- beads_of_interest_short[-grep("ifn|bare|anti|ebna|rabbit|^RBD$", beads_of_interest_short)]
```

### For review: Antibody signals in the different phases, just light grey lines for each subject, and then a trendline (loess smooth) for the overall pattern:

```{r}

## Using Z-scores:
plots_zscore <- lapply(covid_beads, function(bead) {
  ggplot(serology_norm_rm_empty_sinfo, aes(x = sampling_phase, y = scale(.data[[bead]], scale= TRUE, center = TRUE))) +
    geom_line(aes(group = idnr), color = "lightgrey", alpha = 0.3) + 
    labs(title = paste0(bead), x = "Sampling phase") +
    theme_classic() +
    ylab(paste0("Z-score.", bead)) + 
    guides(fill = "none", color = "none") +
    geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "black", formula = 'y ~ x')
})

# Transform data to be relative to Phase 1 per individual
serology_relative <- serology_norm_rm_empty_sinfo %>%
  group_by(idnr) %>%
  mutate(across(all_of(covid_beads), function(x) {
    # Extract the value at Phase 1 for a specific person
    baseline <- x[sampling_phase == "Phase 1"]
    
    # If they are missing Phase 1, or have multiple records, fallback to NA
    if (length(baseline) != 1) {
      return(NA_real_) 
    }
    
    # Otherwise: perform the subtraction
    return(x - baseline)
  })) %>%
  ungroup() %>%
  # Clean up individuals who couldn't be normalized (missing Phase 1)
  filter(!is.na(.data[[covid_beads[1]]]))

plots_relative <- lapply(covid_beads, function(bead) {
  ggplot(serology_relative, aes(x = sampling_phase, y = .data[[bead]])) +
    geom_line(aes(group = idnr), color = "lightgrey", alpha = 0.3) + 
    # Horizontal reference line at 0 to emphasize no-change
    geom_hline(yintercept = 0, linetype = "dashed", color = "darkgrey", alpha = 0.7) +
    labs(title = paste0(bead), x = "Sampling phase") +
    theme_classic() +
    ylab(paste0("Relative change to Phase 1")) + 
    guides(fill = "none", color = "none") +
    # Trendline showing population average change over time
    geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "black", formula = 'y ~ x')
})

```

### Sampling time plots in relation to vaccination (Figure 1b):

```{r}

sample_info_extended <- sample_info |>  left_join(sample_info_wide %>% select(idnr, chc19aldy, qc19_3aldy, IgG_bamse, male, smoke_c19_3, BMI_c19, qc19_3q9_1_yes, qc19_3q9_2_yes, c19_3_postcovid_symp, sminet_date  ), by = c("idnr_pad" = "idnr"))

serology_norm_rm_empty_sinfo <- serology_norm_rm_empty_sinfo |>
  left_join(sample_info_wide %>% select(idnr, chc19aldy, qc19_3aldy, IgG_bamse, male, smoke_c19_3, BMI_c19, qc19_3q9_1_yes, qc19_3q9_2_yes, c19_3_postcovid_symp, sminet_date  ), by = c("idnr_pad" = "idnr"))

ggplot( serology_norm_rm_empty_sinfo, aes( x = as.Date(sampling_time), fill = sampling_phase)) +
  geom_histogram( col = "white") +
  theme_classic() +
  labs(
    x = "Sampling date",
    y = "Number of Samples",
    fill = "Sampling Phase",
    title = "Number of samples per date"
  ) +
  scale_fill_brewer(palette = "Set2")

long_dates <- serology_norm_rm_empty_sinfo |>
  mutate( sampling_time = as.Date(sampling_time),
          dose1 = as.Date(date1),
          dose2 = as.Date(date2),
          dose3 = as.Date(date3),
          sminet_date = as.Date(sminet_date),
          
          
          dose1 = if_else(dose1 <= sampling_time, dose1, as.Date(NA)),
          dose2 = if_else(dose2 <= sampling_time, dose2, as.Date(NA)),
          dose3 = if_else(dose3 <= sampling_time, dose3, as.Date(NA)),
          sminet_date = if_else(sminet_date <= sampling_time, sminet_date, as.Date(NA))
          
  ) |>
  select(sample_name, sampling_time, sampling_phase, dose1, dose2, dose3, sminet_date) |>
  pivot_longer(
    cols = c(sampling_time, dose1, dose2, dose3, sminet_date),
    names_to = "date_type",
    values_to = "date"
  ) |>
  mutate(
    date = as.Date(date)
  )

### Color the phases differentyly:
plot_data <- long_dates |>
  filter(date_type != "sminet_date") |>
  mutate(
    date_type = as.character(date_type),
    fill_group = if_else(
      date_type == "sampling_time",
      sampling_phase,
      date_type
    )
  )

ggplot(plot_data, aes(x = date, fill = fill_group)) +
  geom_histogram(
    col = "white",
    alpha = 0.8,
    position = "stack",
    binwidth = 20
  ) +
  facet_wrap(
    ~ factor(date_type, levels = c("sampling_time", "dose1", "dose2", "dose3")),
    ncol = 1,
    scales = "free_y"
  ) +
  scale_fill_manual(
    values = c(
      "Phase 1" = "#66C2A5",
      "Phase 2" = "#FC8D62",
      "Phase 3" = "#8DA0CB",
      "dose1" = "#aed6f1",
      "dose2" = "#93c47d",
      "dose3" = "#004953"
    )
  ) +
  scale_x_date(
    date_breaks = "2 month",
    date_labels = "%Y-%m"
  ) +
  theme_classic() +
  theme(
    strip.text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  labs(
    title = "Number of samples across different dates",
    x = "Date",
    y = "Number of samples",
    fill = ""
  )

```

### Age distribution per sex (Fig. 1c):

```{r}
# Load age from phase 1:
age_bamse_df <- read.csv("bamse_age_all_phases.csv")

plot_data <- serology_norm_rm_empty_sinfo |>
  select(idnr, male, sampling_phase) |>
  filter( sampling_phase == "Phase 1") |>
  left_join(age_bamse_df, by = "idnr") |>
  distinct(idnr, .keep_all = TRUE) |>  # Keep only unique idnr
  mutate(sex = case_when( 
    male == "1" ~ "Male",
    male == "0" ~ "Female"
  ))

mean_value <- plot_data %>%
  group_by(sex) %>%
  summarize(mean = mean(qc19_1aldy))

median_value <- plot_data %>%
  group_by(sex) %>%
  summarize(median = median(qc19_1aldy))

ggplot(plot_data, aes(x = qc19_1aldy, col = sex)) +
  geom_density(size = 2) +
  theme_classic(base_size = 12) +
  labs(x = "Age", y = "Density", title = "Age (Phase 1) distribution by sex", subtitle = "median = dotted, mean = dashed") +
  scale_color_manual(values = c("Male" = "#697", "Female" = "#eac14a")) +
  geom_vline(data = mean_value, aes(xintercept = mean, color = sex), linetype = "dashed") +
  geom_vline(data = median_value, aes(xintercept = median, color = sex), linetype = "dotted") 
```

### Distribution of vaccination doses per sampling phase (Figure 1d):

```{r}
date_matrix <- serology_norm_rm_empty_sinfo |>
  mutate( dose1 = as.Date(date1),
          dose2 = as.Date(date2),
          dose3 = as.Date(date3),
          sminet_date = as.Date(sminet_date)) |>
  select(sample_name, sampling_time, sampling_phase, dose1, dose2, dose3, sminet_date)


# Categorize samples based on the number of doses at sampling_time
date_matrix <- date_matrix %>%
  mutate(
    dose_count = case_when(
      is.na(dose1) & is.na(dose2) & is.na(dose3) ~ "0 doses",  # No doses at all
      sampling_time < dose1 ~ "0 doses",  
      sampling_time >= dose1 & (sampling_time < dose2 | is.na(dose2)) ~ "1 dose",  
      sampling_time >= dose2 & (sampling_time < dose3 | is.na(dose3)) ~ "2 doses",  
      sampling_time >= dose3 ~ "3 doses",  
      TRUE ~ NA_character_
    )
  )

source("comp_groups.R")

plot_obj <- comp_groups(date_matrix$sampling_phase, date_matrix$dose_count, p_adj_method = "fdr", p_stars = TRUE, discr_palette =  scale_fill_manual(values = c(
  "0 doses" = "#c0c0c0",  # Red
  "1 dose" = "#aed6f1",   # Blue
  "2 doses" = "#93c47d",  # Green
  "3 doses" =  "#004953"   # Purple
)) )
```

### Samling time in relation to infection and distribution of PCR-confirmed infections per sampling phase (Supplementary Figure 1 a and b):

```{r}
plot_data <- long_dates |>
  filter(!(date_type %in% c("dose1", "dose2", "dose3"))) |>
  mutate(
    date_type = as.character(date_type),
    fill_group = if_else(
      date_type == "sampling_time",
      sampling_phase,
      date_type
    )
  )

# Create the plot
plot <- plot_data |> 
  # filter(!(date_type %in% c("dose1", "dose2", "dose3"))) |>
  #  mutate( date_type = as.character(date_type)) |>
  ggplot( aes(x = date,  fill = fill_group)) +
  # Barplot for all date types
  geom_histogram(
    aes(
      fill = ifelse(date_type == "sampling_time", sampling_phase, date_type)
    ),
    col = "white",
    alpha = 0.8,
    position = "stack"
  ) +
  facet_wrap(~ factor(date_type, levels = c("sampling_time", "sminet_date")), ncol = 1, scales = "free_y") +
  theme_classic() +
  labs(
    x = "Date",
    y = "Number of samples",
    fill = "",
    title = "Confirmed infection across different dates"
  ) +
  scale_fill_manual(
    values = c(
      "1" = "#66C2A5",
      "2" = "#FC8D62",
      "3" = "#8DA0CB",
      "sminet_date" = "#c3bfa1"
    )
  ) +
  # Add more x-axis ticks
  scale_x_date(
    date_breaks = "3 month",  
    date_labels = "%Y-%m"    
  ) + 
  theme(
    strip.text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

source("comp_groups.R")

# Categorize samples based on the number of doses at sampling_time
date_matrix <- date_matrix %>%
  mutate(
    dose_count = case_when(
      is.na(dose1) & is.na(dose2) & is.na(dose3) ~ "0 doses",  # No doses at all
      sampling_time < dose1 ~ "0 doses",  
      sampling_time >= dose1 & (sampling_time < dose2 | is.na(dose2)) ~ "1 dose",  
      sampling_time >= dose2 & (sampling_time < dose3 | is.na(dose3)) ~ "2 doses",  
      sampling_time >= dose3 ~ "3 doses",  
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    infected = case_when(
      sampling_time > sminet_date ~ "Infected",
      TRUE ~ "Unknown"
    )
  )

plot_obj <- comp_groups(date_matrix$sampling_phase, date_matrix$infected, p_adj_method = "fdr", p_stars = TRUE, discr_palette =  scale_fill_manual(values = c(
  "Unknown" = "grey88",
  "Infected" = "#c3bfa1"
)) )

plot_obj$plt +   labs(
  x = "Sampling phase",
  y = "% of samples",
  title = "Percentage of samples with PCR-confirmed infection",
  fill = ""
) 

plot + (plot_obj$plt +   labs(
  x = "Sampling phase",
  y = "% of samples",
  title = "Percentage of samples with PCR-confirmed infection",
  fill = ""
) ) +  plot_layout(widths = c(2, 1))

```

# Serology analysis code:

### Hierarchical clustering of samples based on antibody response (Figure 2a):

```{r}
library(fpc)

# Define cluster characteristics:
clust_dist <- "maximum"
clust_method <- "ward.D2"
n_clust <- 4


six_prot_beads <- c("nucleocapsid.full", "nucleocapsid.c", "spike.s1" ,"spike.wt", "ba.2.rbd.omicron", "rbd.wt.sars.2")


## Take out only normalized mfi data:
clust_data <- serology_norm_rm_empty %>%
  select(!grep("_0.4", colnames(serology_norm_rm_empty))) |>
  select( all_of(six_prot_beads))


clust_stab <- T
# Perform stability test
if (clust_stab) {
  boot_clustfun <- function(x, dist_method = clust_dist, cl_method = clust_method, n_cl = n_clust) {
    # x, matrix with data to cluster
    
    clust_results <- dist(x, method = dist_method) %>% hclust(method = cl_method)
    clusts <- cutree(clust_results, n_cl)
    
    out_list <- list("result" = clust_results,
                     "nc" = length(unique(clusts)),
                     "clusterlist" = lapply(sort(unique(clusts)), function(x) {clusts == x}),
                     "partition" = clusts,
                     "clustermethod" = "hclust")
  }
  set.seed(123)
  bs <- clusterboot(clust_data, B = 100, clustermethod = boot_clustfun, count = F)
  #Jacard index - the size of the intersection divided by the size of the union of two label sets, is used to compare set of predicted labels for a sample to the corresponding set of labels in y_true:
  cat("Mean Jaccard indices: ")
  cat(bs$bootmean)
  
  # Get results from first run as clustering to look at
  hcl <- bs$result$result
} else {
  hcl <- clust_data %>% dist(method = clust_dist) %>% hclust(method = clust_method)
}

clusterSamples <- cutree(hcl, k = n_clust)
serology_norm_rm_empty$cluster_maximum_6prot_omi <- as.character(clusterSamples)

### Plot the heatmaps
library(ghibli)
library(pheatmap)
library(viridis)

rownames(clust_data) <- serology_norm_rm_empty$sample_name
rownames(serology_norm_rm_empty) <- serology_norm_rm_empty$sample_name

rownames(serology_norm_rm_empty_sinfo) <- serology_norm_rm_empty_sinfo$sample_name

serology_norm_rm_empty_sinfo$cov_status <- as.factor(serology_norm_rm_empty_sinfo$cov_status)
serology_norm_rm_empty_sinfo$sampling_phase <- as.factor(serology_norm_rm_empty_sinfo$sampling_phase)

annot_colours <- list("sampling_phase" = RColorBrewer::brewer.pal(4, "Set2") %>% setNames(levels(serology_norm_rm_empty_sinfo$sampling_phase)), 
                      "cluster_maximum_6prot_omi" = as.character(ghibli_palettes$LaputaMedium[7:4] ) %>% setNames( levels( as.factor(serology_norm_rm_empty_sinfo$cluster_maximum_6prot_omi))))

# The heatmap (transposed):
heatmap_cluster_t <- pheatmap(t(clust_data), 
                              color = viridis::magma(101),
                              cluster_rows = T, cluster_cols = as.hclust(as.dendrogram(hcl)), clustering_distance_rows = clust_dist,
                              clustering_method = clust_method, show_rownames = T, show_colnames = F, cutree_cols = n_clust,
                              annotation_col = serology_norm_rm_empty_sinfo %>% select(cluster_maximum_6prot_omi, sampling_phase) ,
                              annotation_colors = annot_colours, border_color = NA, main = paste0("Hierarchical clustering, ward.D2, ", clust_dist))


heatmap_cluster_t


#### Plot the Ab intensities for the clusters:
plots <- lapply(six_prot_beads, function(bead) {
  ggplot(serology_norm_rm_empty_sinfo, aes(x = cluster_maximum_6prot_omi, y = .data[[bead]], color = factor(cluster_maximum_6prot_omi))) +
    geom_violin(aes( fill = cluster_maximum_6prot_omi),cex = 1.3, trim = FALSE, alpha = 0.5) +
    geom_boxplot(width = 0.1, position = position_dodge(0.9), col = "black", alpha = 0.5) +
    labs(title = paste0(bead), y = bead, x = "Cluster") +
    theme_classic() +
    #facet_wrap(~sampling_phase) +
    scale_colour_ghibli_d("LaputaMedium", direction = -1) +
    scale_fill_ghibli_d("LaputaMedium", direction = -1) +
    #scale_color_manual( values = c("#D27FBF", "#96E7FF", "#4DAB71", "#FFEC98")) +
    #scale_color_brewer( palette = "Set2") +
    ylab(paste0("normalized log2(mfi).", bead)) + # +
    guides( fill = "none")
})

print(plots)

## Alluvial plot for cluster travel:
library(ggalluvial)
library(ghibli)

alluvial_plot <- serology_norm_rm_empty_sinfo |>
  select( sampling_phase, cluster_maximum_6prot_omi, idnr_pad  ) |>
  rownames_to_column( var = "rownames") |>
  ggplot( aes(alluvium = idnr_pad, x = sampling_phase, stratum = cluster_maximum_6prot_omi)) + 
  #geom_alluvium(color= "grey" , alpha = 0.5) +
  geom_flow(aes(fill = cluster_maximum_6prot_omi), col = "black" , alpha = .5) + 
  geom_stratum( color = "black",aes(fill=cluster_maximum_6prot_omi)) +
  scale_fill_ghibli_d("LaputaMedium", direction = -1)+
  ylab("Number of BAMSE subjects") +
  xlab("") +
  ggtitle("Change in cluster-belonging")+
  theme_classic(15)

print(alluvial_plot)

```

### Correlation between protein variants (Supplementary Figure 2):

```{r}

library(GGally)


## RBD:
serology_norm_rm_empty_sinfo |>
  select(contains("rbd")) |>  # Select columns containing "rbd"
  select(-contains("0.4")) |> # Remove columns containing "0.4"
  select(-"rbd") |>
  select(-contains("nearly")) |>
  select(-contains("sars.1")) |>
  ggpairs(
    upper = list(continuous = wrap("cor", method = "spearman")),
    title = "RBD protein correlation"
  )

## Spike:
serology_norm_rm_empty_sinfo |>
  select(contains("spike")) |>  # Select columns containing "rbd"
  select(-contains("0.4")) |> # Remove columns containing "0.4"
  ggpairs(
    upper = list(continuous = wrap("cor", method = "spearman")),
    title = "Spike protein correlation"
  )

## Nucleocapsid:
serology_norm_rm_empty_sinfo |>
  select(contains("nucle")) |>  # Select columns containing "rbd"
  select(-contains("0.4")) |> # Remove columns containing "0.4"
  ggpairs(
    upper = list(continuous = wrap("cor", method = "spearman")),
    title = "Nucleocapsid protein correlation"
  )
```

### Determine seropositivity using population density cutoffs

Population-based threshold of 6 standard deviations (SD) of the estimated negative proportion above the Gaussian population peak. The negative population consisted of unvaccinated individuals without PCR-confirmed infection in sampling phase 1 & 2.

```{r}
dens_cl <- function(x, s, neg_prop, fullx_phase, method) {
  
  # Exclude controls and pools
  x_pop <- x
  
  # Save a cutoff, population density peak, and plot per protein
  cutoff_vec = peak_vec <- vector("numeric", ncol(x_pop))
  names(cutoff_vec) = names(peak_vec) <- colnames(x_pop)[1:(ncol(x_pop))]
  
  plt_lst <- vector("list", ncol(x_pop))
  names(plt_lst) <- colnames(x_pop)[1:(ncol(x_pop))]
  
  plt_lst2 <- vector("list", ncol(fullx_phase))
  names(plt_lst2) <- colnames(fullx_phase)[1:(ncol(fullx_phase))]
  
  for (prot in colnames(x_pop)[1:(ncol(x_pop))]) {
    val <- x_pop[, prot]
    
    # Get population density peak and distance from peak
    dens <- density(val)
    peak <- dens$x[which.max(dens$y)]
    pdist <- abs(val - peak)
    
    # Get variation of neg_prop around peak
    qtl <- quantile(pdist, probs = neg_prop)
    vtn <- sd(val[pdist <= qtl])
    
    # Classify points
    # Use a cutoff of peak - 6 for selected binders (due to high prevalence instead of low)
    cutoff <- peak + ifelse(grepl("anti.human|ebna", prot, ignore.case = T), -6, 6) * vtn
    
    peak_vec[prot] <- peak
    cutoff_vec[prot] <- cutoff
    
    # Make plot
    p <- ggplot(x_pop %>% dplyr::select(!!prot)) +
      geom_histogram(aes_string(x = prot), size = 1.2, fill = "white", col = "grey30") +
      geom_vline(xintercept = peak, linetype = "dashed") +
      geom_vline(xintercept = cutoff, linetype = "dotted") +
      labs( title = "Phase 1 and 2",
            subtitle = method) +
      theme_classic() +
      theme(axis.line = element_line(size = 1),
            axis.text = element_text(size = 14),
            axis.title = element_text(size = 16))
    
    plt_lst[[prot]] <- p
    
    # Make plot on all sample phases:
    p <- ggplot(fullx_phase %>% dplyr::select(!!prot)) +
      geom_histogram(aes_string(x = prot), size = 1.2, fill = "white", col = "grey30") +
      geom_vline(xintercept = peak, linetype = "dashed") +
      geom_vline(xintercept = cutoff, linetype = "dotted") +
      labs( title = "All phases",
            subtitle = method) +
      theme_classic() +
      theme(axis.line = element_line(size = 1),
            axis.text = element_text(size = 14),
            axis.title = element_text(size = 16))
    
    plt_lst2[[prot]] <- p
    
    
  }
  
  return(list("cutoffs" = cutoff_vec,
              "peaks" = peak_vec,
              "plots_cutoffs" = plt_lst,
              "plots_all_phases" = plt_lst2))
}

# Estimated proportion of positive samples from vaccinations and infections in phase 1 and 2:
serology_norm_rm_empty_sinfo <- serology_norm_rm_empty_sinfo |>
  mutate( cov_status = case_when( is.na(sminet_case) & is.na(vaccdose1) ~ "V-I-",
                                   is.na(sminet_case) & vaccdose1 == "Yes" ~ "V+I-",
                                   sminet_case=="Yes" & is.na(vaccdose1) ~ "V-I+",
                                   sminet_case=="Yes" & vaccdose1 == "Yes" ~ "V+I-",
                                   ))


filtered_data <- serology_norm_rm_empty_sinfo |> 
  filter( sampling_phase %in% c("Phase 1", "Phase 2"))

# Calculate the number of rows matching cov_status conditions
matching_rows <- filtered_data |> 
  filter(cov_status %in% c("V-I+", "V+I-", "V+I+") | IgG_bamse == 1 ) |> 
  nrow()
# Calculate the total number of rows in the filtered dataset
total_rows <- nrow(filtered_data)
# Calculate the proportion
pos_prop <- matching_rows / total_rows

# Make cutoffs for each covid19 protein, INF protein and EBNA1
binfo <- data.frame(beads_of_interest)
colnames(binfo)[1] <- "analyte"
binfo$analyte_number <- 1:72

prot_incl <- binfo[-grep("bare|rabbit", binfo$analyte), 2]
mfi <- filtered_data[,c(2:73)]
sinfo <- filtered_data[,74:ncol(filtered_data)]

method <- "Density peak for P1&2 + 6SD(estimated neg)"
dens_cutoffs <- dens_cl(mfi[, c(prot_incl)], sinfo, 1 - pos_prop, serology_norm_rm_empty_sinfo[,c(2:73)], method)


# Classify all samples (including controls) for each protein
cls <- serology_norm_rm_empty_sinfo[, c("sample_name", names(dens_cutoffs$cutoffs))]
for (i in colnames(cls)[-1]) {
  cls[, i] <- cls[, i] > dens_cutoffs$cutoffs[i]
}

#      Create a summary table with % of TRUE/FALSE for each antibody:

# Reshape the data from wide to long format
long_cls <- cls |>
  pivot_longer(
    cols = -sample_name,   # Exclude the first column (sample_name)
    names_to = "protein",  # Create a 'protein' column with the protein names
    values_to = "value"    # Create a 'value' column with the corresponding TRUE/FALSE values
  )

# Calculate the percentage of TRUE and FALSE per protein
seropos_summary_table <- long_cls |>
  #group_by(protein) |>
  #mutate(total_count = n() ) |>
  #ungroup() |>
  group_by(protein, value) |>               # Group by protein and value (TRUE/FALSE)
  summarise(count = n()) |> # Count the number of TRUE and FALSE values per protein
  mutate(percentage = count / sum(count) * 100)

# Calculate the percentage of TRUE and FALSE per protein and phase
seropos_summary_table_phase <- long_cls |>
  left_join(serology_norm_rm_empty_sinfo |> select( c("sample_name", "sampling_phase")), by = "sample_name") |>
  group_by(sampling_phase, protein, value) |>   # Group by protein and value (TRUE/FALSE)
  summarise(count = n()) |> # Count the number of TRUE and FALSE values per protein
  mutate(percentage = count / sum(count) * 100)

```

### Check how many are considered sero +/- per serocluster:

```{r}

# Create a summary table with % TRUE/FALSE per cluster:

# Initialize an empty list to store the small tables
seropos_cluster_comp <- list()

# Use lapply to create a list of tables, one for each protein
seropos_cluster_comp <- lapply(unique(seropos_summary_table$protein), function(prot) {
  
  # Filter the data for the current protein
  protein_data <- long_cls |>
    left_join(serology_norm_rm_empty_sinfo[, c("sample_name", "sampling_phase", "IgG_bamse", "cluster_maximum_6prot_omi")]) |>
    filter(protein == prot)
  
  # Create a contingency table
  contingency_table <- table(protein_data$cluster_maximum_6prot_omi, protein_data$value)
  
  # Convert to data frame for easier manipulation
  df <- as.data.frame.matrix(contingency_table)
  
  
  return(df)
})

# Assign names to the list elements based on the protein names
names(seropos_cluster_comp) <- unique(seropos_summary_table$protein)

### Write it as a table:
seropositive_clusters_df <- map_dfr(covid_proteins, function(name) {
  # Extract the table for the current protein
  tbl <- seropos_cluster_comp[[name]] %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "Serocluster") # Convert row names to a column
  
  # Ensure both "FALSE" and "TRUE" columns exist, replacing missing ones with 0
  if (!"FALSE" %in% colnames(tbl)) {
    tbl$`FALSE` <- 0
  }
  if (!"TRUE" %in% colnames(tbl)) {
    tbl$`TRUE` <- 0
  }
  
  # Create a data frame with required columns
  tbl %>%
    mutate(
      analyte = name, # Add analyte column
      percent_true = round(`TRUE` / (`FALSE` + `TRUE`) * 100, 1) # Calculate percentage
    ) %>%
    select(analyte, Serocluster, `FALSE`, `TRUE`, percent_true) # Reorder columns
})



### Seroclusters per phase:
# Initialize an empty list to store the small tables
seropos_cluster_comp_perphase <- list()

# Use lapply to create a list of tables, one for each protein
seropos_cluster_comp_perphase <- lapply(unique(covid_proteins), function(prot) {
  
  # Filter the data for the current protein
  protein_data <- long_cls |>
    left_join(serology_norm_rm_empty_sinfo[, c("sample_name", "sampling_phase", "IgG_bamse", "cluster_maximum_6prot_omi")], by = "sample_name") |>
    filter(protein == prot)
  
  # Create a contingency table for sampling_phase & cluster vs. value
  contingency_table <- table(protein_data$sampling_phase, protein_data$cluster_maximum_6prot_omi, protein_data$value)
  
  # Convert the array to a tidy data frame
  df <- as.data.frame.table(contingency_table) |> 
    rename(sampling_phase = Var1, cluster = Var2, value = Var3, count = Freq) |> 
    pivot_wider(names_from = value, values_from = count, values_fill = list(count = 0)) |> 
    rename(`FALSE` = `FALSE`, `TRUE` = `TRUE`)
  
  return(df)
})


names(seropos_cluster_comp_perphase) <- unique(covid_proteins)

### Write it as a table:
seropositive_clusters_phase_df <- map_dfr(covid_proteins, function(name) {
  # Extract the table for the current protein
  tbl <- seropos_cluster_comp_perphase[[name]] %>% 
    as.data.frame() #%>% 
  #rownames_to_column(var = "Serocluster") # Convert row names to a column
  
  # Ensure both "FALSE" and "TRUE" columns exist, replacing missing ones with 0
  if (!"FALSE" %in% colnames(tbl)) {
    tbl$`FALSE` <- 0
  }
  if (!"TRUE" %in% colnames(tbl)) {
    tbl$`TRUE` <- 0
  }
  
  # Create a data frame with required columns
  tbl %>%
    mutate(
      analyte = name, # Add analyte column
      percent_true = round(`TRUE` / (`FALSE` + `TRUE`) * 100, 1) # Calculate percentage
    ) %>%
    select(analyte, sampling_phase, cluster, `FALSE`, `TRUE`, percent_true) # Reorder columns
})


```

### Agreement across RBD variants (Supplementary Figure 3):

Let the RBD WT variant be 100%, and then the percentage of positive for the other variants, sort in decreasing order.

```{r}
rbd_variants <- colnames(cls)[grep("rbd", colnames(cls))]
rbd_variants <- rbd_variants[-grep("^rbd$|0.4", rbd_variants)]

reference_col <- "rbd.wt.sars.2"  # Define reference variant

rbd_agreement_df <- cls |>
  select( sample_name, all_of(rbd_variants)) |>
  left_join(
    serology_norm_rm_empty_sinfo |>
      select(sample_name, sampling_phase, idnr),
    by = "sample_name"
  ) 

# Count number of samples positive for reference
total_reference_positive <- sum(rbd_agreement_df[[reference_col]] == "TRUE", na.rm = TRUE)

# Compute agreement percentage for each variant
agreement_results <- lapply(rbd_variants, function(variant) {
  # Count samples where reference is TRUE and variant is also TRUE
  agreement_count <- sum(rbd_agreement_df[[variant]] == "TRUE" & rbd_agreement_df[[reference_col]] == "TRUE", na.rm = TRUE)
  
  # Compute percentage agreement
  percentage_agreement <- (agreement_count / total_reference_positive) * 100
  
  return(data.frame(variant = variant, percentage = percentage_agreement))
}) |> 
  bind_rows()


# Step 3: Sort variants by agreement percentage (highest to lowest)
agreement_results <- agreement_results |> arrange(desc(percentage))

# Step 4: Visualize as a bar plot
agreement_results |> filter(! variant == "rbd.nearly.omicron") |>
  ggplot( aes(x = reorder(variant, -percentage), y = percentage)) +
  geom_bar(stat = "identity", width = 1, show.legend = FALSE, alpha = 0.6, col = "white", fill = "grey", size=2) +
  theme_classic(base_size = 12) +
  labs(title = "Agreement of RBD Variants with rbd.wt",
       x = "RBD Variant",
       y = "Agreement Percentage (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline( yintercept = 75, linetype = "dashed")

```

### Agreement between Luminex data (DBS) and ELISA (Plasma) (Supplementary Figure 4):

```{r}

# Create a summary table for phase 2 with how many are positive in this assay and how many are in the ELISA:
# Initialize an empty list to store the small tables
seropos_elisa_comp <- list()

# Use lapply to create a list of tables, one for each protein
seropos_elisa_comp <- lapply(unique(seropos_summary_table$protein), function(prot) {
  
  # Filter the data for the current protein
  protein_data <- long_cls |>
    left_join(serology_norm_rm_empty_sinfo[, c("sample_name", "sampling_phase", "IgG_bamse", "cluster_maximum_6prot_omi")]) |>
    filter( sampling_phase == "Phase 2")  |>
    filter(protein == prot)
  
  # Create a contingency table
  contingency_table <- table(protein_data$IgG_bamse, protein_data$value)
  
  # Convert to data frame for easier manipulation
  df <- as.data.frame.matrix(contingency_table)
  
  
  return(df)
})

# Assign names to the list elements based on the protein names
names(seropos_elisa_comp) <- unique(seropos_summary_table$protein)

## Save it to a df:
seropositive_elisa_df <- map_dfr(names(seropos_elisa_comp), function(name) {
  # Extract the table for the current protein
  tbl <- seropos_elisa_comp[[name]] %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "ELISA_IgG_pos") # Convert row names to a column
  
  
# Ensure both "FALSE" and "TRUE" columns exist, replacing missing ones with 0
  if (!"FALSE" %in% colnames(tbl)) {
    tbl$`FALSE` <- 0
  }
  if (!"TRUE" %in% colnames(tbl)) {
    tbl$`TRUE` <- 0
  }
  
  # Create a data frame with required columns
  tbl %>%
    mutate(
      analyte = name, # Add analyte column
      percent_true = round(`TRUE` / (`FALSE` + `TRUE`) * 100, 1) # Calculate percentage
    ) %>%
    select(analyte, ELISA_IgG_pos, `FALSE`, `TRUE`, percent_true) # Reorder columns
})

cohen_kappa_df <- NULL
for (protein in covid_beads){
  df_temp <- seropositive_elisa_df |>
    filter( analyte == protein) |>
    select(ELISA_IgG_pos,  `FALSE`, `TRUE`) |>
    mutate( ELISA_IgG_pos = case_when( ELISA_IgG_pos == 0 ~ "FALSE",
                                       ELISA_IgG_pos == 1 ~ "TRUE")) |>
    column_to_rownames("ELISA_IgG_pos") |>
    as.matrix()
  
  cohen_kappa_protein <- psych::cohen.kappa(df_temp)
  cohen_kappa_df <- rbind( cohen_kappa_df, c(protein, cohen_kappa_protein$kappa))
  
}

## Create a file that contains seropositivity for spike proteins and their ELISA_igG:
filtered_cls_df <- cls %>%
  left_join(
    serology_norm_rm_empty_sinfo %>% select(sample_name, sampling_phase, IgG_bamse),
    by = "sample_name"
  ) %>%
  filter(sampling_phase == "Phase 2") %>%
  select(sample_name, sampling_phase, IgG_bamse, contains("spike"))

#Pivot longer for analysis
long_cls_elisa_df <- filtered_cls_df %>%
  pivot_longer(cols = contains("spike"), names_to = "spike_feature", values_to = "spike_value")

summary_df <- long_cls_elisa_df %>%
  group_by(spike_feature, spike_value, IgG_bamse) %>%
  summarise(n = n(), .groups = "drop")

# Create the complete grid of all combinations
all_combos <- expand_grid(
  spike_feature = unique(summary_df$spike_feature),
  spike_value = c(TRUE, FALSE),
  IgG_bamse = c(0, 1)
)

# Join with actual counts and fill missing combos with 0
summary_df_plot <- all_combos %>%
  left_join(summary_df, by = c("spike_feature", "spike_value", "IgG_bamse")) %>%
  mutate(
    n = replace_na(n, 0),
    spike_value = factor(spike_value, levels = c(FALSE, TRUE)),
    IgG_bamse = factor(IgG_bamse, levels = c(0, 1)),
    # Reverse the facet order
    spike_feature = factor(spike_feature, levels = rev(unique(summary_df$spike_feature)))
  )

plot1 <- summary_df_plot |>
  filter( !grepl("0.4", spike_feature )) |>
  filter( !is.na(IgG_bamse)) |>
  ggplot( aes(x = IgG_bamse, y = spike_value, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 4) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  facet_wrap(~ spike_feature) +
  labs(
    x = "ELISA",
    y = "Luminex assay",
    fill = "Count",
    title = "Seropositivity for Spike vs ELISA measured IgG"
  ) +
  theme_classic(base_size = 12)

```

### Antibody level response to vaccination and infection (Supplementary Figure 5):

```{r}

### Antibody levels against time since infection (Days since PCR positive)
sminet_infected_subjects <- serology_norm_rm_empty_sinfo |> filter(sminet_case == "Yes") |> pull(idnr) |> unique()

plot2 <-  serology_norm_rm_empty_sinfo %>% 
  filter(idnr %in% sminet_infected_subjects) %>%
  mutate(
    time_since_sminet = as.Date(sampling_time) - as.Date(sminet_date),
    time_since_sminet_group = case_when(
      time_since_sminet < 0 ~ "pre",
       time_since_sminet >= 0 & time_since_sminet <= 30 ~ "0-30",
       time_since_sminet > 5 & time_since_sminet <= 10 ~ "6-10",
       time_since_sminet > 30 ~ "30+"
      #time_since_sminet >= 0 ~ "post",
    ),
    time_since_sminet_group = factor(time_since_sminet_group, levels = c("pre", "0-30",  "30+"))
    #time_since_sminet_group = factor(time_since_sminet_group, levels = c("pre", "post"))
  ) %>%
  filter(Time_Sample_Dose1 < 0) %>%
  group_by(idnr) %>%
  filter(!all(time_since_sminet_group == "pre")) |>
  pivot_longer(cols = all_of(beads_of_interest_short), names_to = "bead", values_to = "mfi") %>%
  filter(bead %in% c("nucleocapsid.full", "spike.wt", "rbd.wt.sars.2" )) %>%
  ggplot(aes(x = time_since_sminet, y = mfi, fill = time_since_sminet_group)) +
  geom_vline( xintercept = 0, linetype = "dotted", col = "grey40") +
  geom_vline( xintercept = 30, linetype = "dotted", col = "grey40") +
  geom_line(aes(group = idnr), color = "gray80", alpha = 0.2) +
  geom_point(size = 3, shape = 21, col = "white" ) +
  theme_classic() +
  labs( y = "log2(MFI) [AU]", x = "Days since PCR+") +
   scale_fill_manual( values = c("grey", "#e93e3a", "#fdc70c"), name = "Days since PCR+") +
  #scale_fill_manual( values = c("grey",  "#fdc70c"), name = "+") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~bead)


### Antibody levels against time since vaccination
vaccinated_subjects <- serology_norm_rm_empty_sinfo |> filter(vaccdose1 == "Yes") |> pull(idnr) |> unique()
nucleocapsid_positive_samples <- long_cls |> filter(protein == "nucleocapsid.full" & value == "TRUE") |> pull(sample_name) ## samples taken after infection

serology_vaccination_data <- serology_norm_rm_empty_sinfo |>
  #filter(vaccdose1 == "Yes" & is.na(sminet_case)) |>
  filter( idnr %in% vaccinated_subjects) |> 
  filter(!sample_name %in% nucleocapsid_positive_samples) |>
  mutate(
    sampling_time = as.Date(sampling_time),
    dose1_date = as.Date(date1),
    dose2_date = as.Date(date2),
    dose3_date = as.Date(date3)
  ) |>
  rowwise() |>  # Process each row individually
  mutate(
    # Combine dose dates into a list and filter those before sampling_time
    dose_dates = list(na.omit(c(dose1_date, dose2_date, dose3_date))),
    dose_dates_before_sampling = list(dose_dates[dose_dates < sampling_time]),
    
    # Identify the closest dose date before the sampling time
    closest_dose_date = max(dose_dates_before_sampling, na.rm = TRUE),
    
    # Calculate time since the closest dose in months
    time_since_any_dose = time_length(interval(closest_dose_date, sampling_time), "days"),
    
    # Add the time to vaccine dose 1 to the samples that were taken before vaccination:
    time_since_any_dose = case_when( time_since_any_dose == "Inf" ~ time_length(interval(dose1_date, sampling_time), "days"),
                                     T ~ time_since_any_dose),
    
    
    # Determine which dose was closest
    closest_dose = case_when(
      closest_dose_date == dose1_date ~ "dose1",
      closest_dose_date == dose2_date ~ "dose2",
      closest_dose_date == dose3_date ~ "dose3",
      TRUE ~ "No dose"
    )
  ) |>
  ungroup()

serology_vaccination_data <- serology_vaccination_data |>
  mutate( time_since_vaccination_group = case_when(
    time_since_any_dose < 0 ~ "pre",
    time_since_any_dose >= 0 & time_since_any_dose <= 5 ~ "0-5",
    time_since_any_dose > 5 & time_since_any_dose <= 10 ~ "6-10",
     time_since_any_dose > 10 & time_since_any_dose <= 30~ "11-30",
     time_since_any_dose > 30 ~ "30+"
    #time_since_any_dose >= 0 ~ "post"
  ),
   time_since_vaccination_group = factor(time_since_vaccination_group, levels = c("pre", "0-5",  "6-10", "11-30", "30+"))
  #time_since_vaccination_group = factor(time_since_vaccination_group, levels = c("pre", "post"))
) 

plot <-  serology_vaccination_data |> 
  filter(!all(time_since_vaccination_group == "pre")) |>
  pivot_longer(cols = all_of(beads_of_interest_short), names_to = "bead", values_to = "mfi") %>%
  filter(bead %in% c("nucleocapsid.full", "spike.wt", "rbd.wt.sars.2" )) %>%
  filter(!is.na(time_since_vaccination_group)) |>
  ggplot(aes(x = time_since_any_dose, y = mfi, fill = time_since_vaccination_group)) +
  geom_vline( xintercept = 0, linetype = "dotted", col = "grey40") +
   geom_vline( xintercept = 30, linetype = "dotted", col = "grey40") +
  geom_line(aes(group = idnr), color = "gray80", alpha = 0.2) +
  geom_point(size = 3, shape = 21, col = "white" ) +
  theme_classic() +
  #stat_compare_means(label = "p.signif", method = "t.test", ref.group = "pre") +
  labs(y = "log2(MFI) [AU]", x = "Days since latest dose") +
  #scale_fill_manual( values = c("grey",  "#fdc70c"), name = "") +
  #scale_color_brewer( palette = "Set2")
  #scale_fill_manual( values = c("grey", "#e93e3a", "#ed683c", "#f3903f", "#fdc70c"), name = "Days since latest dose") +
  scale_fill_manual( values = c("grey", "#e93e3a","#FE8D01", "#FFC78F" , "#fdc70c"), name = "Days since latest dose") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap( ~ bead) 

plot2 / plot + plot_annotation(tag_levels = 'A')

## Stability of Ab levels since vaccination, no infection (based on sminet case and if they are TRUE against nucleocapsid.full for serology test (Figure S5 c & d)

protein_to_plot <- c("nucleocapsid.full", "spike.s1s2.foldon")

serology_norm_rm_empty_sinfo |>
  filter(vaccdose1 == "Yes" &  is.na(sminet_case)) |>
  filter( ! sample_name %in% nucleocapsid_positive_samples ) |>
  mutate(
    sampling_time = as.Date(sampling_time),
    dose1_date = as.Date(date1),
    time_since_dose1_months = time_length(interval(dose1_date, sampling_time), "months")
  ) |>
  #select( sample_name, idnr, all_of(five_prot_beads), sampling_phase, time_since_dose1_months, nr_vacc_doses ) |>
  select( sample_name, idnr, all_of(protein_to_plot), sampling_phase, time_since_dose1_months, nr_vacc_doses ) |>
  pivot_longer( cols=!c(sample_name, idnr, sampling_phase, time_since_dose1_months, nr_vacc_doses ), values_to = "norm.mfi", names_to = "analyte") |>
  ggplot( aes( x = time_since_dose1_months, y = norm.mfi )) +
  #geom_line(aes(group = idnr), alpha = 0.5) +
  geom_point( aes(col = as.factor(nr_vacc_doses) ), alpha = 0.9, size = 2) +
  facet_wrap(~analyte) +
  geom_smooth(method = 'loess', col = "#E9C46A") + 
  theme_classic(13) +
  labs( title= "Time since vaccine dose 1", subtitle = "(excl. sminet infected and nucleocapsid.full positive)", y = "Normalized log2 MFI", x = "Months since Vaccine dose 1") +
  #scale_color_manual(values= c("#D81B60", "#1E88E5", "#004D40"))
  scale_color_manual(values= c("#aed6f1", "#93c47d", "#004953"))



# Figure S5e
# Assign colors per analyte
library(ggnewscale)


# Extract cutoff values from the data frame
cutoff_values <- data.frame(
  analyte = names(dens_cutoffs$cutoffs[c(nuc_protein, s_protein)]),  # Protein names
  cutoff = dens_cutoffs$cutoffs[c(nuc_protein, s_protein)]  # Corresponding cutoff values
)

cutoff_values <- cutoff_values |>   mutate( family = case_when( analyte %in% nuc_protein ~ "N",
                                                                analyte %in% s_protein ~ "S"))

protein_colors <- c(
  "nucleocapsid_acro" = pal_fill[3], 
  "nucleocapsid.c" = pal_fill[4], 
  "nucleocapsid.full" = pal_fill[5],  # Ensure nucleocapsid.full is in green shades
  "spike.s1" = "#fea214", 
  "spike.s1s2.foldon" = "#f78e50", 
  "spike.wt" = "#ec7b73"
)

# Family colors for smooth curves and horizontal lines
#family_colors <- c("N" = pal_fill[4], "S" = "#f6bc67")
family_colors <- c("N" = "#08589E", "S" = "#f99d48")



serology_norm_rm_empty_sinfo |>
  #left_join(sample_info_wide[ , c("idnr", "sminet_date")], by = "idnr") |>
  filter(sminet_case == "Yes" & is.na(vaccdose1)) |>
  mutate(
    sampling_time = as.Date(sampling_time),
    sminet_date = as.Date(sminet_date),
    time_since_sminet = as.numeric(sampling_time - sminet_date),
    time_since_sminet_months = time_length(interval(sminet_date, sampling_time), "months")
  ) |>
  select(
    sample_name, idnr, sampling_phase, time_since_sminet_months, all_of(c(nuc_protein, s_protein))
  ) |>
  pivot_longer(cols = !c(sample_name, idnr, sampling_phase, time_since_sminet_months), 
               values_to = "norm.mfi", names_to = "analyte") |>
  mutate(
    family = case_when(
      analyte %in% nuc_protein ~ "N",
      analyte %in% s_protein ~ "S"
    )
  ) |>
  ggplot(aes(x = time_since_sminet_months, y = norm.mfi)) +
  # POINTS - Use analyte colors
  geom_point(aes(color = analyte), alpha = 0.8, size = 2) +
  scale_color_manual(
    name = "Analyte", 
    values = protein_colors  # Define colors for points (analyte)
  ) +
  new_scale_colour() +
  # SMOOTH LINES - Use family colors
  geom_smooth(method = 'loess', aes(color = family, fill = family), formula = y ~ x) +  
  theme_classic(12) +
  # TWO DIFFERENT COLOR MAPPINGS
  scale_color_manual(
    name = "Family", 
    values = family_colors  # Define colors for points (analyte)
  ) +
  scale_fill_manual(
    name = "Family", 
    values = family_colors   # Define colors for smooth lines (family)
  ) +
  # HORIZONTAL LINES - Use viral protein family colors
  geom_hline(
    data = cutoff_values, 
    aes(yintercept = cutoff, linetype = analyte, color = family)  
  ) +
  labs(
    title = "Anti-N and S signal after infection",
    subtitle = "Exclude vaccinated subjects",
    y = "Normalized log2 MFI",
    x = "Months since confirmed infection"
  )

```

# Proteomics analysis code:

### Linear models protein \~ traits (BMI, sex, age, smoking):

```{r}

library(lmerTest)

model_vars <- c("chc19aldy", "male", "smoke_c19_3", "BMI_c19" )

olink_proteins_oi <- colnames(olink_bamse_protpqn_blank_rm_intnorm)[c(2:(ncol(olink_bamse_protpqn_blank_rm_intnorm))-1)]

# Initialize a data frame to store the results
results_lm_p2 <- data.frame(protein = character(), var = character(), adj_R2 = numeric(), intercept = numeric(), slope = numeric(),  p_value_intercept = numeric(),  p_value_slope = numeric(), stringsAsFactors = FALSE)

# Loop over each protein and each model variable
for (protein in olink_proteins_oi) {
  for (var in model_vars) {
    
    formula <- as.formula(paste(protein, "~", var))
    
    phase2_data <- olink_bamse_protpqn_blank_rm_intnorm_si_ex |> filter(sampling_phase == "Phase 2")
    
    # Fit the linear mixed model
    model <- tryCatch(
      lm(formula, data = phase2_data),
      error = function(e) NULL
    )
    
    adjr2 <- summary(lm(formula, data = phase2_data))$adj.r.squared
    
    estimates <- if (!is.null(model)) {
      summary(model)$coefficients[, "Estimate"]
    } else {
      NA
    }
    
    p_values <- if (!is.null(model)) {
      summary(model)$coefficients[, "Pr(>|t|)"]
    } else {
      NA
    }
    
    
    results_lm_p2 <- rbind(results_lm_p2, data.frame(protein = protein, var = var,  adj_R2 = adjr2, intercept = estimates[1], slope = estimates[2] ,  p_value_intercept = p_values[1], p_value_slope =  p_values[2] ))
    
    
  }
  
}

# FDR adjust data:
results_lm_p2 <- results_lm_p2 |> 
  group_by(var) |> 
  mutate(fdr_slope = p.adjust(p_value_slope, method = "fdr")) |> 
  ungroup()

write.csv(results_lm_p2, "lm phase 2 against information_protpqn intesnity data_fdr across variable_250928.csv")


## Plot fdr vs slope:
plot_list <- lapply(model_vars, function(variable) {
  dat <- results_lm_p2_in |>
    filter(var == variable) |>
    mutate(
      rank_sig = min_rank(fdr_slope),
      label = case_when(
        rank_sig <= 10 ~ protein,
        TRUE ~ NA_character_
      ),
      significant_color = case_when(fdr_slope < 0.05 ~ "sig",
                                    TRUE ~ "not sig"),
      max_abs_slope = max(abs(slope), na.rm = TRUE)
    )
  
  p <- ggplot(dat, aes(x = slope, y = -log10(fdr_slope), col = significant_color)) +
    geom_point(alpha = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    geom_blank(aes(x = max_abs_slope)) +
    geom_blank(aes(x = -max_abs_slope)) +
    theme_classic(10) +
    scale_color_manual(values = c("not sig" = "grey79", "sig" = "black")) +
    ylim(0, 40) +
    labs( title = paste0(variable)) +
    theme(plot.title = element_text(hjust = 0.5))
  
  # Only add geom_text if there is at least one "sig"
  if (any(dat$significant_color == "sig")) {
    p <- p + geom_text(
      data = dat,
      aes(label = label),
      size = 3,
      color = "black",
      nudge_y = 0.15,
      check_overlap = TRUE
    )
  }
  
  return(p)
})

plot_list[[4]] + plot_list[[1]] + plot_list[[2]] + plot_list[[3]] + plot_layout(guides = "collect")

```

### PCA plot of proteomics data and correlation between PC1-3 and clinical variables (Fig S7):

```{r}

olink_bamse_protpqn_blank_rm_intnorm_si <- olink_bamse_protpqn_blank_rm_intnorm |>
  left_join(samp_info_key, by = c("sample_id" = "sampleid")) |>
  left_join(sample_info, by = c("samplename" = "sample_name"))

### Correlate all proteins as well to PC1, PC2 and PC3 together with the clinical variables:
pca_in <- olink_bamse_protpqn_blank_rm_intnorm_si |>
  dplyr::select( samplename, all_of(proteins_of_interest)) |>
  column_to_rownames( var = "samplename")
pca_in <- pca_in[ complete.cases( pca_in),]

pca_result <- prcomp(pca_in[, 1:ncol(pca_in)])
pc_scores <- as.data.frame(pca_result$x)


pc_scores <- pc_scores %>%
  mutate(samplename = rownames(pc_scores)) %>%
  left_join(olink_bamse_protpqn_blank_rm_intnorm_si |> dplyr::select( samplename, sample_date, nr_vacc_doses, sampling_phase, cluster_maximum_6prot_omi ), by = "samplename") |>
  left_join( olink_bamse_protpqn_blank_rm_intnorm_si_ex |> dplyr::select( samplename,  male, smoke_c19_3, BMI_c19,qc19_3aldy),  by = "samplename") |>
  left_join( serology_norm_rm_empty_sinfo |> dplyr::select(sample_name, nucleocapsid.full, spike.wt), by = c("samplename" = "sample_name")) |>
  mutate( nr_vacc_doses = case_when( is.na(nr_vacc_doses) ~ 0,
                                     T ~nr_vacc_doses)) |>
  #smoke_c19_3 = case_when( is.na(smoke_c19_3) ~ 0,
  #                         T ~ smoke_c19_3)) |>
  mutate( elution_date = as.Date("2022-09-30"),
          sample_date = as.Date(sample_date),
          storage_time = time_length(interval(sample_date, elution_date), "months"),
          sample_date_nr = as.numeric(sample_date),
          serocluster_numeric = as.numeric(cluster_maximum_6prot_omi),
          sampling_phase_nr = case_when( sampling_phase == "Phase 1" ~ 1,
                                         sampling_phase == "Phase 2" ~ 2,
                                         sampling_phase == "Phase 3" ~ 3)) |>
  left_join(olink_bamse_protpqn_blank_rm_intnorm_si |> dplyr::select( samplename, all_of(proteins_of_interest)), by = "samplename") |>
  rowwise() %>% 
  mutate(mean_npx = mean(c_across(all_of(proteins_of_interest)), na.rm = TRUE)) %>%
  ungroup()


clinical_vars <- c("sample_date_nr", "nr_vacc_doses", "sampling_phase_nr", "serocluster_numeric", "nucleocapsid.full", "spike.wt", "male", "BMI_c19", "qc19_3aldy", "storage_time", "mean_npx" )

# For selected PCs (e.g., top 5), correlate each with clinical vars
library(corrr)

cor_results <- correlate(pc_scores[, c("PC1", "PC2", "PC3", clinical_vars, proteins_of_interest)], method = "spearman")

cor_long <- cor_results %>%
  pivot_longer(cols = -term, names_to = "PC", values_to = "correlation") %>%
  filter(!term %in% c("PC1", "PC2", "PC3", "PC4", "PC5"))  # remove PC self-correlations

cor_order <- cor_long |>
  filter(PC == "PC1") |>
  arrange(desc(correlation)) |>
  dplyr::select(term, order_pc1 = correlation)


## only clinical variables:
cor_long |>
  filter(PC %in% c("PC1", "PC2", "PC3")) |>
  filter( term %in% clinical_vars) |>
  #left_join(cor_order, by = "term") |>
  #mutate(term = forcats::fct_reorder(term, order_pc1, .desc = TRUE)) |>
  ggplot(aes(y = PC, x = term, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limit = c(-1, 1), space = "Lab",
    name = "Spearman\nCorrelation"
  ) +
  geom_text(aes(label = round(correlation, 2)), size = 2) +
  theme_minimal() +
  labs(
    y = "Principal Components", x = "Clinical variables",
    title = "Correlation between PCA components and clinical variables"
  ) +
  #theme(axis.text.x = element_text(angle = 45, hjust = 1))
  theme(
    axis.text.x = element_text(
      angle = 60,           # more angled labels
      hjust = 1,            # right-align
      size = 7              # smaller x-axis labels
    ),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 10)
  )


### Spearman correlations but getting p-values for correlations too:
pcs <- c("PC1", "PC2", "PC3")
vars <- c(clinical_vars, proteins_of_interest)


cor_results_p <- expand_grid(
  pc = pcs,
  var = vars
) %>%
  mutate(
    test = map2(pc, var, ~ {
      x <- pc_scores[[.x]]
      y <- pc_scores[[.y]]
      if (sum(!is.na(x) & !is.na(y)) >= 3) {
        cor.test(x, y, method = "spearman")
      } else {
        NA
      }
    }),
    rho = map_dbl(test, ~ ifelse(is.list(.x), .x$estimate, NA_real_)),
    p_value = map_dbl(test, ~ ifelse(is.list(.x), .x$p.value, NA_real_))
  ) %>%
  dplyr::select(-test) %>%
  mutate(fdr = p.adjust(p_value, method = "fdr"))

cor_long |> filter(PC == "PC1") |> filter(term %in% proteins_of_interest) |> ggplot( aes( x = correlation )) + geom_histogram( stat = "bin", color = "black", fill = "white") +
  theme_classic() + 
  labs( title = "Protein correlation to PC1", x = "Spearman's rho") +
  scale_x_continuous(limits = c(-1, 1)) +
  geom_vline( xintercept = 0, linetype = "dashed", color = "grey", linewidth = 1.2) +
  geom_vline( xintercept = 0.5, linetype = "dotted", color = "orange", linewidth = 1.2) +
  geom_vline( xintercept = -0.5, linetype = "dotted", color = "orange", linewidth = 1.2)

```

### Comparison of protein-trait associations in BAMSE and UK biobank (UKB) (Fig S8):

```{r}

source("bamse_vs_ukb.R")

ukb1 <- read.csv("ukb1.csv") # table 5
ukb2 <- read.csv("ukb2.csv") # table 7

results_lm_p2_in <- read.csv("lm phase 2 against information_protpqn intesnity data_fdr across variable_250928.csv")

## Adjust the names in ukb1 and ukb2 so that these are the same as in bamse file:
ukb1$assay[grep("HLA-E", ukb1$assay)] <- "HLA.E"
ukb2$assay[grep("HLA-E", ukb2$assay)] <- "HLA.E"

ukb1$assay[grep("HLA-DRA", ukb1$assay)] <- "HLA.DRA"
ukb2$assay[grep("HLA-DRA", ukb2$assay)] <- "HLA.DRA"

ukb_vs_bamse_result <- bamse_vs_ukb(results_lm_p2_in, ukb1, ukb2)

# Whick proteins are not present in both data sets?
setdiff(results_lm_p2_in$protein, ukb1$assay)

ukb_vs_bamse_result$plot

write.csv(ukb_vs_bamse_result$df, "ukb vs bamse slopes results_p value_251001.csv")
```

### **Comparison of protein-trait associations in BAMSE and population survey. (Fig. S9):**

```{r}

### bamse_vs_vaccserol function:

vaccserol_beta <- read_excel("table_S7_vaccserol.xlsx")

## Adjust the names in ukb1 and ukb2 so that these are the same as in bamse file:
ukb1$assay[grep("HLA-E", ukb1$assay)] <- "HLA.E"
ukb2$assay[grep("HLA-E", ukb2$assay)] <- "HLA.E"

ukb1$assay[grep("HLA-DRA", ukb1$assay)] <- "HLA.DRA"
ukb2$assay[grep("HLA-DRA", ukb2$assay)] <- "HLA.DRA"

vaccserol_vs_bamse_result <- bamse_vs_vaccserol(results_lm_p2_in, vaccserol_beta)

vaccserol_vs_bamse_result$plot + labs( x = "Beta from vaccination serology", y = "Beta form BAMSE")

write.csv(vaccserol_vs_bamse_result$df, "vaccination serology vs bamse slopes results_p value_251024.csv")
```

### Changes of protein levels in relation to storage time (Fig. S10):

```{r}
## LMM against storage time, adjust for sex and do this only for the serocluster 1 samples

## Take out the serocluster 1 samples that remain in serocluster 1 all phases:
serocluster1_3p <- serology_norm_rm_empty_sinfo |>
  group_by(idnr) |>
  filter(all( cluster_maximum_6prot_omi == 1)) |>
  pull(sample_name)

#library(lme4)
library(lmerTest)
# Initialize an empty data frame to store the results
lm_results_storagetime <- data.frame(protein = character(),
                                     slope = numeric(),
                                     p_value = numeric(),
                                     stringsAsFactors = FALSE)

olink_proteins <- olink_proteins_oi


plot_list <- list()
# Loop over each protein
for (protein in olink_proteins) {
  
  # Perform the data selection and linear modeling
  data <- olink_bamse_protpqn_blank_rm_intnorm |>
    dplyr::rename(sampleid = sample_id) |> 
    # select(!sample_id) |>
    left_join(samp_info_key, by = "sampleid") |>
    left_join(sample_info, by = c("samplename" = "sample_name")) |>
    mutate(elution_date = as.Date("2022-09-30"),
           sampling_time = as.Date(sampling_time),
           storage_time = time_length(interval(sampling_time, elution_date), "months"),
           sampling_phase = plot_variable) |>
    left_join( sample_info_wide |> dplyr::select(idnr, male), by = "idnr") |>
    mutate(sex = case_when( male == 0 ~ "female",
                            male == 1 ~ "male")) |>
    filter( samplename %in% serocluster1_3p)
  #filter(!is.na(idnr))
  
  
  # Fit a linear mixed model
  lmer_fit <- lmerTest::lmer(data[[protein]] ~ storage_time + (1 | idnr) + sex, data = data)
  
  # Extract fixed effects
  lmer_summary <- summary(lmer_fit)$coefficients
  
  # Extract slope and p-value for storage_timeserocluster
  slope <- lmer_summary["storage_time", "Estimate"]
  p_value <- lmer_summary["storage_time", "Pr(>|t|)"]
  
  # Store the results
  lm_results_storagetime <- rbind(lm_results_storagetime, data.frame(protein = protein,
                                                                     slope = slope,
                                                                     p_value = p_value,
                                                                     stringsAsFactors = FALSE))

  
}


lm_results_storagetime$fdr_value <- p.adjust( lm_results_storagetime$p_value , method = "fdr")
lm_results_storagetime$bonferroni <- p.adjust( lm_results_storagetime$p_value , method = "bonferroni")

lm_results_storagetime <- lm_results_storagetime |>
  mutate( fdr_color = case_when( fdr_value < 0.01 &  slope < 0 ~ "sig.down",
                                 fdr_value < 0.01 &  slope > 0 ~ "sig.up",
                                 T ~ "not sig")) |>
  mutate( bonferroni_color = case_when( bonferroni < 0.01 &  slope < 0 ~ "sig.down",
                                        bonferroni < 0.01 &  slope > 0 ~ "sig.up",
                                        T ~ "not sig")) 

# Select the top 10 rows with the lowest bonferroni values
top10_proteins <- lm_results_storagetime %>%
  arrange(bonferroni) %>%
  head(10)

up_proteins <- lm_results_storagetime |>
  filter(bonferroni_color == "sig.up")

library(ggrepel)

label_proteins <- rbind(top10_proteins, up_proteins)

ggplot(lm_results_storagetime, aes( x = slope, y = -log10(bonferroni))) +
  geom_point( aes(col = bonferroni_color ), size = 3) +
  geom_hline(yintercept = -log10(0.01), linetype="dashed", 
             color = "black", size = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", 
             color = "black", size = 1) +
  scale_color_manual( values = c("grey", "#A6CEE3" , "#1F78B4")) +
  xlim(-0.14, 0.14) +
  labs( title = "Slopes for Protein signal vs storage time (months)", subtitle = "LMM + Bonferroni adjusted p-value") +
  theme_classic(10) +
  geom_text_repel(data = label_proteins, 
                  aes(label = protein), 
                  max.overlaps = Inf, # Ensures all labels are shown
                  box.padding = 0.5, 
                  point.padding = 0.3, 
                  segment.color = "grey50") 

# Count how many bonferroni values are below 0.01
count_below_0_01 <- lm_results %>%
  filter(bonferroni < 0.01) %>%
  tally()

unstable_proteins <- lm_results_storagetime %>%
  filter(bonferroni < 0.01) |> pull(protein)

# Get the top 10 proteins with the lowest bonferroni values
top_10_proteins <- lm_results %>%
  arrange(bonferroni) %>%  # Sort by bonferroni in ascending order
  slice_head(n = 10) %>%   # Select the top 10 rows
  select(protein, bonferroni)  # Select only the protein and bonferroni columns

## Plot distribution plot of the slopes:
plot1 <- ggplot(lm_results_storagetime, aes( x = slope, fill = bonferroni_color )) +
  geom_histogram(  col = "black", alpha = 0.8) +
  theme_classic(10) +
  geom_vline(aes(xintercept= 0 ),
             color="black", linetype="dashed", size=1) + 
  labs( title = "Distribution of slopes for Protein signal vs storage time (months)", subtitle = "Slopes from LMM") +
  scale_fill_manual( values = c("grey", "#A6CEE3" , "#1F78B4")) +
  theme_classic(10)

###                   Fig S10 c:
hpa_file <- read.delim("./datafiles/proteinatlas.tsv")

hpa_stor_df <- storage_time_effect %>%
  mutate(sig_storage = case_when( bonferroni < 0.01 ~ "sig",
                                  T ~ "not sig")) |>
  left_join(olink_uniprot, by = c("protein" = "Assay"))

hpa_stor_df$UniProt[grep("HLA.E", hpa_stor_df$protein)] <- "P13747"
hpa_stor_df$UniProt[grep("HLA.DRA", hpa_stor_df$protein)] <- "P01903"
hpa_stor_df$UniProt[grep("SIT1", hpa_stor_df$protein)] <- "Q9Y3P8"


hpa_file$Uniprot[grep("RABGAP1L", hpa_file$Gene)] <- "Q5R372"
hpa_file$Uniprot[grep("PNLIPRP2", hpa_file$Gene)] <- "P54317"
hpa_file$Uniprot[grep("SIRPB1", hpa_file$Gene)] <- "O00241"
hpa_file$Uniprot[grep("CDSN", hpa_file$Gene)] <- "Q15517"
hpa_file$Uniprot[grep("TGFB1", hpa_file$Gene)] <- "P01137"

hpa_stor_df <- hpa_stor_df |>
  left_join(hpa_file , by = c("UniProt" = "Uniprot" ))

hpa_stor_df_v2 <- hpa_stor_df %>%
  mutate(
    Predicted_secreted = str_detect(
      Protein.class %||% "",
      regex("Predicted secreted proteins", ignore_case = TRUE)
    ),
    Predicted_intracellular = str_detect(
      Protein.class %||% "",
      regex("Predicted intracellular proteins", ignore_case = TRUE)
    ),
    Predicted_membrane = str_detect(
      Protein.class %||% "",
      regex("Predicted membrane proteins", ignore_case = TRUE)
    ),
    
    # Combination category
    class_combo = case_when(
      Predicted_secreted & Predicted_intracellular ~ "Secreted + Intracellular",
      Predicted_secreted & Predicted_membrane ~ "Secreted + Membrane",
      Predicted_intracellular & Predicted_membrane ~ "Intracellular + Membrane",
      Predicted_secreted &  Predicted_membrane & Predicted_intracellular ~ "Secreted + Membrane + Intracellular",
      Predicted_secreted ~ "Secreted only",
      Predicted_intracellular ~ "Intracellular only",
      Predicted_membrane ~ "Membrane only",
      TRUE ~ "Other / none"
    )
  )


source("comp_groups.R")

plot1 <- comp_groups( hpa_stor_df_v2$class_combo, hpa_stor_df$sig_storage,  p_adj_method = "fdr", p_stars = TRUE, x_name = "Protein class", g_name = "Significant storage effect" )

plot1$plt
```

### **Spearman and intraclass correlation across repeated samplings** (Fig. S17):

**Calculate Inter-class correlations. ICC \> 0.8 are considered as stable:**

The Intraclass correlation is used as a measure of association when studying the reliability of raters.

ICC (close to 1) indicates high similarity between rater's scores.

ICC (close to 0) indicates rater's scores are not similar.

Usage:

```         
ICC(x,missing=TRUE,alpha=.05,lmer=TRUE,check.keys=FALSE)
```

```{r}
library(psych)

#------------------------------ Correlation and ICC per protein:
#Spearman correlation:
corr_matrix_list <- lapply(olink_proteins_oi, function(protein) {
  
  
  # Extract data for the given protein
  protein_data <-  olink_bamse_protpqn_blank_rm_intnorm_si |>
    #select(samplename, idnr, sampling_phase, cluster_maximum_6prot_omi, !!sym(protein))
  select(samplename, idnr, sampling_phase, !!sym(protein))
  
  # Split data by phase
  phase1 <- protein_data %>% filter(sampling_phase == "Phase 1")
  phase2 <- protein_data %>% filter(sampling_phase == "Phase 2")
  phase3 <- protein_data %>% filter(sampling_phase == "Phase 3")
  
  # Merge phase data on sample names
  phase1_vs_phase2 <- merge(phase1, phase2, by = "idnr", suffixes = c("_P1", "_P2"))
  phase2_vs_phase3 <- merge(phase2, phase3, by = "idnr", suffixes = c("_P2", "_P3"))
  phase1_vs_phase3 <- merge(phase1, phase3, by = "idnr", suffixes = c("_P1", "_P3")) 
  
  # Calculate Spearman's rho for Phase 1 vs Phase 2
  rho_phase1_vs_phase2 <- cor(phase1_vs_phase2[[paste0(protein, "_P1")]], 
                              phase1_vs_phase2[[paste0(protein, "_P2")]], 
                              method = "spearman", 
                              use = "complete.obs")
  
  
  # Calculate Spearman's rho for Phase 2 vs Phase 3
  rho_phase2_vs_phase3 <- cor(phase2_vs_phase3[[paste0(protein, "_P2")]], 
                              phase2_vs_phase3[[paste0(protein, "_P3")]], 
                              method = "spearman", 
                              use = "complete.obs")
  
  rho_phase1_vs_phase3 <- cor(phase1_vs_phase3[[paste0(protein, "_P1")]], 
                              phase1_vs_phase3[[paste0(protein, "_P3")]], 
                              method = "spearman", 
                              use = "complete.obs")
  
  
  # Return the correlation values
  return(c(IDnr = protein, corr_v1v2 = rho_phase1_vs_phase2, corr_v2v3 = rho_phase2_vs_phase3, corr_v1v3 = rho_phase1_vs_phase3))
  
})


# Convert list to data frame
corr_matrix_list_df <- do.call(rbind, corr_matrix_list)

colnames(corr_matrix_list_df) <- c("protein", "phase1_vs_phase2", "phase2_vs_phase3", "phase1_vs_phase3" )

corr_matrix_list_df <- as.data.frame(corr_matrix_list_df)
corr_matrix_list_df[,2:4] <- apply(corr_matrix_list_df[,2:4], 2, as.numeric)

corr_plot_protein_intnorm <- corr_matrix_list_df |>
  pivot_longer( cols = contains("phase" ), names_to = "Comparison", values_to = "spearman" ) |>
  ggplot( aes( x = spearman, col = Comparison, fill = Comparison)) +
  geom_density( alpha = 0.5) +
  theme_classic(11) +
  scale_fill_brewer( palette = "Pastel2") +
  scale_color_brewer( palette = "Pastel2") +
  labs(title = "Protein correlation", x = "Spearman's Rho") +
  theme( # remove the vertical grid lines
    panel.grid.major.y = element_blank() ,
    # explicitly set the horizontal lines (or they will disappear too)
    panel.grid.major.x = element_line( size=.1, color="black", linetype = "dotted" ) 
  ) +
  xlim(-1, 1)

##                      ICC (Supp Fig S17b)
icc_matrix_list <- lapply(olink_proteins_oi, function(protein) {
  
  
  # Extract data for the given protein
  protein_data <- olink_bamse_protpqn_blank_rm_intnorm_si |>
    dplyr::select(samplename, idnr, sampling_phase, !!sym(protein))
  
  # Split data by phase
  phase1 <- protein_data %>% filter(sampling_phase == "Phase 1")
  phase2 <- protein_data %>% filter(sampling_phase == "Phase 2")
  phase3 <- protein_data %>% filter(sampling_phase == "Phase 3")
  
  # Merge phase data on sample names
  phase1_vs_phase2 <- merge(phase1, phase2, by = "idnr", suffixes = c("_P1", "_P2"))
  
  phase2_vs_phase3 <- merge(phase2, phase3, by = "idnr", suffixes = c("_P2", "_P3"))
  
  phase1_vs_phase3 <- merge(phase1, phase3, by = "idnr", suffixes = c("_P1", "_P3"))
  
  # Calculate ICC for Phase 1 vs Phase 2
  icc_model <-  psych::ICC(cbind(phase1_vs_phase2[[paste0(protein, "_P1")]],
                                 phase1_vs_phase2[[paste0(protein, "_P2")]]), alpha = 0.05, lmer = TRUE, check.keys = TRUE)
  
  icc3_v12 <- icc_model$results$ICC[3]
  
  # Calculate ICC for Phase 2 vs Phase 3
  icc_model <-  psych::ICC(cbind(phase2_vs_phase3[[paste0(protein, "_P2")]],
                                 phase2_vs_phase3[[paste0(protein, "_P3")]]), alpha = 0.05, lmer = TRUE, check.keys = TRUE)
  
  icc3_v23 <- icc_model$results$ICC[3]
  
  # Calculate ICC for Phase 1 vs Phase 3
  icc_model <-  psych::ICC(cbind(phase1_vs_phase3[[paste0(protein, "_P1")]],
                                 phase1_vs_phase3[[paste0(protein, "_P3")]]), alpha = 0.05, lmer = TRUE, check.keys = TRUE)
  
  icc3_v13 <- icc_model$results$ICC[3]
  
  
  # Return the correlation values
  return(c(protname = protein, corr_v1v2 = icc3_v12, corr_v2v3 = icc3_v23, corr_v1v3 =  icc3_v13 ))
  # wrap_plots(plot1 + plot2 + plot_layout(guides = "collect"))
  
})


# Convert list to data frame
icc_matrix_list_df <- do.call(rbind, icc_matrix_list)

colnames(icc_matrix_list_df) <- c("protein", "phase1_vs_phase2", "phase2_vs_phase3", "phase1_vs_phase3" )

icc_matrix_list_df <- as.data.frame(icc_matrix_list_df)

icc_matrix_list_df[,2:4] <- apply(icc_matrix_list_df[,2:4], 2, as.numeric)

icc_plot_protein_intnorm <- icc_matrix_list_df |>
  pivot_longer( cols = contains("phase" ), names_to = "Comparison", values_to = "icc3" ) |>
  ggplot( aes( x = icc3, col = Comparison, fill = Comparison)) +
  geom_density( alpha = 0.5) +
  theme_classic(11) +
  scale_fill_brewer( palette = "Pastel2") +
  scale_color_brewer( palette = "Pastel2") +
  labs(title = "Protein intraclass correlation (ICC3)", x = "Intraclass correlation") +
  xlim(-1, 1) +
  theme( # remove the vertical grid lines
    panel.grid.major.y = element_blank() ,
    # explicitly set the horizontal lines (or they will disappear too)
    panel.grid.major.x = element_line( size=.1, color="black", linetype = "dotted" ) 
  ) +
  xlim(-1, 1)

#                                Plot the proteins that have the highest and lowest ICC:
colnames(icc_matrix_list_df) <- c("protein", "icc_v1v2", "icc_v2v3", "icc_v1v3" )


# Top 5 highest and lowest for icc_v1v2
top_5_high_icc_v1v2 <- icc_matrix_list_df %>%
  arrange(desc(icc_v1v2)) %>%
  head(5)

top_5_low_icc_v1v2 <- icc_matrix_list_df %>%
  arrange(icc_v1v2) %>%
  head(5)

# Top 5 highest and lowest for icc_v2v3
top_5_high_icc_v2v3 <- icc_matrix_list_df %>%
  arrange(desc(icc_v2v3)) %>%
  head(5)

top_5_low_icc_v2v3 <- icc_matrix_list_df %>%
  arrange(icc_v2v3) %>%
  head(5)

# Combine all results into a list
top_5_summary <- list(
  top_5_high_icc_v1v2 = top_5_high_icc_v1v2,
  top_5_low_icc_v1v2 = top_5_low_icc_v1v2,
  top_5_high_icc_v2v3 = top_5_high_icc_v2v3,
  top_5_low_icc_v2v3 = top_5_low_icc_v2v3
)

# Combine them into a single data frame:
top_5_summary_df <- bind_rows(
  top_5_high_icc_v1v2 %>% mutate(category = "top_5_high_icc_v1v2"),
  top_5_low_icc_v1v2 %>% mutate(category = "top_5_low_icc_v1v2"),
  top_5_high_icc_v2v3 %>% mutate(category = "top_5_high_icc_v2v3"),
  top_5_low_icc_v2v3 %>% mutate(category = "top_5_low_icc_v2v3")
)

top_and_bottom <- unique(top_5_summary_df$protein)

plot_list <- lapply(top_and_bottom, function(protein) {
  
  ggplot_data <- olink_bamse_protpqn_blank_rm_intnorm_si %>%
    select(!!sym(protein), sampling_phase, samplename, sample_date, idnr_pad)
  
  print(paste0("protein: " , protein, "n samples: " , nrow(ggplot_data), "n subjects", ggplot_data |> pull(idnr_pad) |> unique() |> length()))
  
  category <- top_5_summary_df[grep(protein, top_5_summary_df$protein), "category"]
  ggplot_data$category <- category
  
  p <- ggplot(ggplot_data, aes(x = sampling_phase, y = !!sym(protein), color = category)) +
    geom_point( size = 0.7) +
    geom_line(aes(group = idnr_pad), alpha = 0.15) +
    theme_classic(10) +
    labs(title = paste0(protein), y = "ProtPQN NPX [AU]", x = "") +
    scale_color_manual(
      values = c(
        "top_5_high_icc_v1v2" = "skyblue2",
        "top_5_high_icc_v2v3" = "skyblue2",
        "top_5_low_icc_v1v2" = "sienna2",
        "top_5_low_icc_v2v3" = "sienna2"
      )
    ) +
    guides( color = "none")
  
  return(p)
})


###---------------------------- Correlate samples from same subjects:

# Compare random vs same subject correlation
set.seed(123)  # Ensure reproducibility

big_df <- olink_bamse_protpqn_blank_rm_intnorm_si

# Randomly select 100 unique idnr values
random_idnrs <- big_df |> distinct(idnr) |> sample_n(807)

# Filter dataset to only keep selected samples (one per idnr)
small_df <- big_df |> filter(idnr %in% random_idnrs$idnr) |> column_to_rownames(var = "samplename") |> select(all_of(olink_proteins_oi)) |> as.matrix()

# Data for correlation and plotting
# Correlation matrix between phases
cor_mat <- cor(t(small_df), method = "spearman")

# Keep only the upper triangle (since the correlation matrix is symmetric and diagonal contains 1s)
cor_mat[lower.tri(cor_mat, diag = T)] <- NA
# Convert to longer format, filter out NAs
cor_mat <- as.data.frame(cor_mat) %>% rownames_to_column("sample1") %>%
  pivot_longer(cols = -sample1, names_to = "sample2", values_to = "corr") %>%
  filter(!is.na(corr))

cor_mat <- cor_mat |>
  mutate(
    num1 = sub("^[A-Z]\\d_", "", sample1),  # Extract number from sample1
    num2 = sub("^[A-Z]\\d_", "", sample2),  # Extract number from sample2
    same_subject = ifelse(num1 == num2, "Same subject", "Different subject")  # Check if they match
  )

# Plot density of correlations
p1 <- ggplot(cor_mat, aes(x = corr, fill = same_subject, color = same_subject)) +
  geom_density(alpha = 0.4) +
  theme_classic() +
  #scale_fill_manual( values = c("#A6CEE3", "#1F78B4" )) + 
  #scale_color_manual( values = c("#A6CEE3", "#1F78B4" )) + 
  scale_fill_brewer( palette = "Pastel2") +
  scale_color_brewer( palette = "Pastel2", ) +
  labs(title = "Sample correlations", x = "Spearman's Rho") +
  theme( # remove the vertical grid lines
    panel.grid.major.y = element_blank() ,
    # explicitly set the horizontal lines (or they will disappear too)
    panel.grid.major.x = element_line( size=.1, color="black", linetype = "dotted" ) 
  ) +
  xlim(-1, 1)

##### Compare correlation for subjects who change serocluster vs the ones who remain in cluster 1
bamse_subject <- unique(olink_bamse_protpqn_blank_rm_intnorm_si$idnr)

corr_matrix_list <- lapply(bamse_subject, function(subject) {
  
  both <- olink_bamse_protpqn_blank_rm_intnorm_si |>
    filter( idnr == subject ) |>
    select( -platepos) |>
    pivot_longer(cols = matches(olink_proteins_oi), names_to = "assay", values_to = "npx")
  
  #phases_for_samp <- unique(both$sampling_phase)
  phases_for_samp <- unique(both$sampling_phase)
  if( "Phase 1" %in%  phases_for_samp && "Phase 2" %in% phases_for_samp) {
    phase1_data <- both %>%
      filter(sampling_phase == "Phase 1") %>%
      select(idnr, assay, npx) %>%
      rename(npx_phase1 = npx)
    
    phase2_data <- both %>%
      filter(sampling_phase == "Phase 2") %>%
      select(idnr, assay, npx) %>%
      rename(npx_phase2 = npx)
    
    combined <- phase1_data %>%
      inner_join(phase2_data, by = c("idnr", "assay"))
    
    print(subject)
    corr_value_v12 <- cor(combined$npx_phase1, combined$npx_phase2, method = "spearman", use = "pairwise.complete.obs")
    
  } else {
    # plot1 <- ggplot( x = 1, y = 2) + geom_point()
    
    corr_value_v12 <- NA
  }
  
  if( "Phase 2" %in%  phases_for_samp && "Phase 3" %in% phases_for_samp) {
    phase2_data <- both %>%
      filter(sampling_phase == "Phase 2") %>%
      select(idnr, assay, npx) %>%
      rename(npx_phase2 = npx)
    
    phase3_data <- both %>%
      filter(sampling_phase == "Phase 3") %>%
      select(idnr, assay, npx) %>%
      rename(npx_phase3 = npx)
    
    
    combined <- phase2_data %>%
      inner_join(phase3_data, by = c("idnr", "assay"))
    
    print(subject)
    corr_value_v23 <- cor(combined$npx_phase2, combined$npx_phase3, method = "spearman", use = "pairwise.complete.obs")
    
  } else {
    # plot2 <- ggplot( x = 1, y = 2) + geom_point()
    corr_value_v23 <- NA
  }
  
  
  if( "Phase 1" %in%  phases_for_samp && "Phase 3" %in% phases_for_samp) {
    phase1_data <- both %>%
      filter(sampling_phase == "Phase 1") %>%
      select(idnr, assay, npx) %>%
      rename(npx_phase1 = npx)
    
    phase3_data <- both %>%
      filter(sampling_phase == "Phase 3") %>%
      select(idnr, assay, npx) %>%
      rename(npx_phase3 = npx)
    
    
    combined <- phase1_data %>%
      inner_join(phase3_data, by = c("idnr", "assay"))
    
    print(subject)
    corr_value_v13 <- cor(combined$npx_phase1, combined$npx_phase3, method = "spearman", use = "pairwise.complete.obs")
    
  } else {
    corr_value_v13 <- NA
  }
  
  
  corr_matrix <-  c(subject, corr_value_v12, corr_value_v23, corr_value_v13)
  
  corr_matrix
  
})


# Convert list to data frame
corr_matrix_list_df <- do.call(rbind, corr_matrix_list)

colnames(corr_matrix_list_df) <- c("idnr", "phase1_vs_phase2", "phase2_vs_phase3", "phase1_vs_phase3" )

corr_matrix_list_df <- as.data.frame(corr_matrix_list_df)


## Mark the ones that are from supernegative subjects:
serocluster1_3p_idnr <- serology_norm_rm_empty_sinfo |>
  group_by(idnr) |>
  filter(all( cluster_maximum_6prot_omi == 1)) |>
  pull(idnr)

corr_plot_sample_intnorm_dual <- corr_matrix_list_df |>
  pivot_longer( cols = contains("phase" ), names_to = "Comparison", values_to = "spearman" ) |>
  mutate( superneg = case_when( idnr %in% serocluster1_3p_idnr ~ "Cluster 1 in all phases",
                                T ~ "Changing cluster")) |>
  ggplot( aes( x = spearman, col = Comparison, fill = Comparison)) +
  geom_density( alpha = 0.3) +
  theme_classic(11) +
  scale_fill_brewer( palette = "Pastel2") +
  scale_color_brewer( palette = "Pastel2") +
  labs(title = "Sample correlation", x = "Spearman's Rho") +
  theme( # remove the vertical grid lines
    panel.grid.major.y = element_blank() ,
    # explicitly set the horizontal lines (or they will disappear too)
    panel.grid.major.x = element_line( size=.1, color="black", linetype = "dotted" ) 
  ) +
  xlim(-1, 1) +
  facet_wrap( ~ superneg,  ncol=1)


#plot <- ggplot()
#plot_build <- ggplot_build(plot)
#plot_build@data

```

## Comparing protein levels in different cluster groups using linear models:

Preparing data frame:

```{r}
##--------------------------------------------------------------------------------------------------------
####                                  Preparing the data frames:
##--------------------------------------------------------------------------------------------------------

bamse_for_lm <- olink_bamse_protpqn_blank_rm_intnorm_si %>%
  filter(length(idnr) > 1, .by = idnr) %>%
  mutate(clust_traj12 = paste0(cluster_maximum_6prot_omi[sampling_phase == "Phase 1"],
                               cluster_maximum_6prot_omi[sampling_phase == "Phase 2"]),
         clust_traj23 = paste0(cluster_maximum_6prot_omi[sampling_phase == "Phase 2"],
                               cluster_maximum_6prot_omi[sampling_phase == "Phase 3"]),
         .by = idnr) %>%
  left_join(olink_bamse_protpqn_blank_rm_intnorm_si_ex %>% dplyr::select(sample_id, male), by = "sample_id")

phase_count12 <- bamse_for_lm %>% distinct(idnr, clust_traj12) %>%
  count(clust_traj12)
phase_count23 <- bamse_for_lm %>% distinct(idnr, clust_traj23) %>%
  count(clust_traj23)

olink_bamse_protpqn_blank_rm_intnorm_si_long <- olink_bamse_protpqn_blank_rm_intnorm_si |>
  pivot_longer(
    cols = all_of(proteins_of_interest),   # Exclude the first column (sample_name)
    names_to = "assay",  # Create a 'protein' column with the protein names
    values_to = "npx"    # Create a 'value' column with the corresponding TRUE/FALSE values
  ) 

### Adjust data for storage time:
time_adjusted_data <- olink_bamse_protpqn_blank_rm_intnorm_si_long |> 
  select(sampling_time, assay, npx, idnr, samplename) |> 
  mutate(elution_date = as.Date("2022-09-30"),
         sampling_time = as.Date(sampling_time),
         storage_time = time_length(interval(sampling_time, elution_date), "months")) |> 
  # filter( assay %in% c("PAPPA", "MVK", "PARP1")) |>
  filter(!assay %in% c("DNPH1", "GLOD4", "PKLR")) |> 
  group_by(assay) |> nest() |>
  mutate(adj_data = map(data, \(x) {
    lmer_result <- resid(lmer(npx ~ storage_time + (1|idnr), data = x[complete.cases(x), ]))
    data.frame(samplename = x$samplename, npx = lmer_result)
  })) |> 
  select(-data) |>
  unnest(adj_data) |> 
  pivot_wider(id_cols = samplename, names_from = assay, values_from = npx) |> 
  left_join(olink_bamse_protpqn_blank_rm_intnorm_si |> 
              select(samplename , cluster_maximum_6prot_omi),
            by = "samplename")


time_adjusted_data_for_lm <- time_adjusted_data |>
  left_join( bamse_for_lm |> dplyr::select(samplename, sample_id, sampling_phase, idnr, male, clust_traj12, clust_traj23  ), by = "samplename")


## Olink proteins in time adjusted data:
olink_proteins_ta <- olink_proteins[-grep("PKLR|GLOD4|DNPH1", olink_proteins)]

```

Function for lm analysis comparing the change in protein levels between different groups:

```{r}
#### Function for performing the lm analysis:
library(ggh4x)
library(ggrepel)

# linear analysis with possibility to do delta delta
lm_analysis <- function(data_in, sample_id = "sampleid", predictor, covar = NULL, stratify_by = NULL,
                        difference_col = NULL, individual_id = NULL, differences_between = NULL) {
  # data_in, list containing data frames with the protein measurements in wide format (named `values`) and sample information in wide format (named `sinfo`). A common sample identifier should exist in both
  # predictor, name of column that should be the main predictor. Must exist in `sinfo` in `data_in` and be numeric.
  # sample_id, name of column in both `values` and `sinfo` that matches the rows of the two data frames
  # covar, vector of column names in `sinfo` that should be included as covariates.
  # stratify_by, name of column to stratify the analysis by (default is NULL, no stratification)
  # difference_col, if differences should be used as the outcome, the name of the column containing the values to calculate difference between. E.g. if two samples come from the same individual at two time points, `difference_col` may be a time point column. Default is NULL (no difference calculation)
  # individual_id, name of column containing individual identifiers. The differences in `difference_col` will be calculated per `individual_id`. Default is NULL (no difference calculation)
  # differences_between, vector of two values existing in `difference_col` to calculate the difference in protein levels between. Default is NULL (no difference calculation)
  
  diff_null <- c(is.null(difference_col), is.null(individual_id), is.null(differences_between))
  if (!sum(diff_null) %in% c(0, 3)) {
    stop("'difference_col', 'individual_id', and 'differences_between' all need to be specified or all NULL.")
  }
  
  # Make formula for linear model
  lm_formula <- paste0("value ~ ", predictor,
                       # Add covariates
                       ifelse(!is.null(covar), " + ", ""),
                       paste0(covar, collapse = " + "))
  
  lm_analysis <- data_in$values %>%
    # Add sample info (columns with covariates, stratification variable, difference variable, IDs)
    left_join(data_in$sinfo %>%
                dplyr::select(all_of(c(sample_id, predictor, covar, difference_col, stratify_by, individual_id))),
              by = sample_id) %>%
    (\(x) {
      if (!is.null(individual_id)) {filter(x, !is.na(get(individual_id)))} else {x}
    })() %>%
    # Convert for long format for analysis
    pivot_longer(cols = -all_of(c(sample_id, predictor, covar, difference_col, stratify_by, individual_id)),
                 names_to = "assay", values_to = "value") %>%
    # Calculate difference in values if columns are provided
    (\(x) {
      if (!is.null(difference_col)) {
        # Perform calculation per individual (individual_id) and protein (assay)
        x %>% group_by(across(all_of(individual_id)), assay) %>%
          # Calculate difference between second and first value of differences_between
          mutate(value = value[get(difference_col) == differences_between[2]] -
                   value[get(difference_col) == differences_between[1]]) %>%
          ungroup() %>%
          # Keep only unique differences for each individual (previous step keeps all rows, leaving repeat values)
          distinct(across(all_of(c(individual_id, "assay"))), .keep_all = T)
      }
      else {x}
    }) %>%
    # Perform analysis per protein and stratification group
    group_by(across(all_of(c("assay", stratify_by)))) %>% nest() %>%
    mutate(median_diff = map(data, \(x) median(x$value[x[[predictor]] == sort(unique(x[[predictor]]))[[2]]]) -
                               median(x$value[x[[predictor]] == sort(unique(x[[predictor]]))[[1]]])),
           median_grp1 = map(data, \(x) median(x$value[x[[predictor]] == sort(unique(x[[predictor]]))[[1]]])),
           median_grp2 = map(data, \(x) median(x$value[x[[predictor]] == sort(unique(x[[predictor]]))[[2]]])),
           lm_res = map(data, \(x) lm(formula = lm_formula, data = x) %>% summary())) %>%
    mutate(# Pick out beta and p-value
      estimate = map(lm_res, \(x) x$coefficients %>% as.data.frame() %>%
                       rownames_to_column("rn") %>% filter(str_detect(rn, predictor)) %>%
                       pull(Estimate)),
      p_value = map(lm_res, \(x) x$coefficients%>% as.data.frame() %>%
                      rownames_to_column("rn") %>% filter(str_detect(rn, predictor)) %>%
                      pull(`Pr(>|t|)`))) %>%
    ungroup() %>%
    dplyr::select(assay, !!stratify_by, median_diff, median_grp1, median_grp2, estimate, p_value) %>%
    unnest(c(median_diff, median_grp1, median_grp2, estimate, p_value)) %>%
    # Calculate FDR
    mutate(fdr = p.adjust(p_value, method = "fdr"), .by = !!stratify_by)
  
  grp1 <- sort(unique(data_in$sinfo[[predictor]]))[[1]]
  grp2 <- sort(unique(data_in$sinfo[[predictor]]))[[2]]
  
  return(list(results = lm_analysis, formula = lm_formula, predictor = predictor,
              covariates = covar, computed_differences = ifelse(is.null(difference_col), F, T),
              difference_params = list(difference_column = difference_col,
                                       differences_between = differences_between,
                                       individual_id = individual_id),
              stratify_by = stratify_by,
              groups = c(grp1, grp2)))
}

# plotting in volcano and mirrored manhattan plot
plot_lm_results <- function(results, colr_by = NA, colr_scale = NULL, sep_by_colr = T,
                            ns_grey = T, signif_threshold = 0.05) {
  
  # Stratified analysis
  strat_log <- !is.null(results$stratify_by)
  strat_name <- results$stratify_by
  
  # Make custom colour scale if non-significant values should be coloured differently
  if (is.null(colr_scale)) {
    if (!all(is.na(colr_by))) {
      # Values to colour by are provided
      
      if (ns_grey) {
        # Introduce a grey "not significant" category
        colr_scale <- scale_colour_manual(
          values = c(brewer.pal(length(unique(colr_by)), "Pastel1"), "grey90") %>%
            setNames(c(as.character(sort(unique(colr_by))), "ns"))
        )
      } else {
        colr_scale <- scale_colour_brewer(palette = "Pastel1")
      }
    } else {
      # No values to colour by are provided
      
      if (ns_grey) {
        # Significant points black, non-significant points grey
        colr_scale <- scale_colour_manual(values = c("Significant" = "black", "ns" = "grey90"))
      }
    }
  }
  
  plt_dat <- results$results %>%
    # Add colouring categories
    (\(x) {
      if (!all(is.na(colr_by))) mutate(x, colr = colr_by)
      else mutate(x, colr = case_when(fdr < signif_threshold ~ "Significant", T ~ "ns"))
    }) %>%
    # Prepare for stratified data
    (\(x) {
      if (strat_log) mutate(x, assay = paste0(assay, "_", get(strat_name)))
      else x
    })()
  
  if (!all(is.na(colr_by))) {
    # Order proteins by colour
    plt_dat <- arrange(plt_dat, colr, assay) %>% mutate(assay = as.integer(factor(assay, levels = assay)))
  }
  
  # Get values for vertical lines to separate colours
  vlines <- plt_dat %>% count(colr) %>% mutate(vline = cumsum(n) + 0.5) %>% dplyr::slice(-nrow(.))
  
  # Change colour name to 'ns' if not significant (if desired)
  if (ns_grey) {
    plt_dat <- mutate(plt_dat,
                      # Keep the original colour names if needed
                      colr_og = colr,
                      colr = case_when(fdr < signif_threshold ~ colr, T ~ "ns"),
                      colr = factor(colr, levels = c(setdiff(as.character(sort(unique(colr))), "ns"), "ns"))) %>%
      # Change order of rows to ensure that ns points are underneath other points
      arrange(desc(colr))
  }
  
  # Mirrored Manhattan plot of results
  mmplt <- plt_dat %>%
    ggplot(aes(x = assay, y = median_diff, size = -log10(fdr),
               colour = if (!all(is.na(colr_by)) | ns_grey) colr else NULL))
  if (sep_by_colr) {mmplt <- mmplt +
    geom_vline(xintercept = vlines$vline, colour = "grey", linetype = "dashed")}
  mmplt <- mmplt +
    geom_point() +
    geom_hline(yintercept = 0, colour = "black", linetype = "dashed") +
    colr_scale +
    # Make y-axis limits symmetric around 0
    scale_y_continuous(limits = center_limits()) +
    labs(y = "∆Change", x = NULL, colour = "Panel", size = "-log10(FDR)") +
    theme_classic(14) +
    theme(axis.text.x = element_blank(),
          axis.line.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          legend.position = "bottom") +
    guides(colour = guide_legend(override.aes = list(size = 3)))
  if (strat_log) {
    mmplt <- mmplt + facet_wrap(as.formula(paste0("~ ", strat_name)), ncol = 1)
  }
  
  # Volcano plot
  vplt <- plt_dat %>%
    ggplot(aes(x = median_diff, y = -log10(fdr), colour = colr)) +
    geom_point(size = 2.5, show.legend = F) +
    geom_hline(yintercept = -log10(signif_threshold), linetype = "dashed") +
    colr_scale +
    scale_x_continuous(limits = center_limits()) +
    theme_classic() +
    geom_text_repel(
      data = subset(plt_dat, fdr < 0.05),
      aes(label = str_remove(assay, "_Phase \\d")),  # Use the `protein` column as the label
      color = "black",
      box.padding = 0.2,
      max.overlaps = Inf
    ) +
    geom_vline( xintercept = 0, linetype = "dashed", col = "grey") #+
  #labs( title = paste("Comparing ", comparison))
  # Faceting based on panels and stratification (Volcano on top of each ohter)
  # if (strat_log | !all(is.na(colr_by))) {
  #   vplt <- vplt + facet_grid(if (strat_log) {rows = vars(get(strat_name))} else {rows = NULL},
  #                             if (sep_by_colr) {cols = case_when(ns_grey ~ vars(colr_og), T ~ vars(colr))}
  #                             else {cols = NULL},
  #                             axes = "all_x")
  # }
  # Faceting based on panels and stratification (Volcano next to each other)
  if (strat_log || !all(is.na(colr_by))) {
    vplt <- vplt + facet_grid(cols = if (strat_log) {vars(get(strat_name))} else {NULL},
                              rows = if (sep_by_colr) {case_when(ns_grey ~ vars(colr_og), T ~ vars(colr))}
                              else {NULL}, axes = "all_y")
  }
  
  rm(list = setdiff(ls(), c("mmplt", "vplt", "colr_by", "ns_grey", "strat_log", "strat_name")))
  
  return(list(mm = mmplt, volcano = vplt))
}

```

### Comparing protein levels in seronegative versus infected subjects (Fig. S18):

```{r}

### Comparing change in protein levels in phase 1 and phase 2 between seronegative and infected:
library(dplyr)
library(purrr)
library(progress)

# PARAMETERS
n_runs <- 100
#min_percentage <- 0.70   # final required frequency
fdr_cutoff <- 0.05
seed_base <- 123

# Precompute group sizes
n12 <- time_adjusted_data_for_lm %>%
  filter(sampling_phase != "Phase 3", clust_traj12 == "12") %>%
  distinct(idnr) %>% nrow()

# Get random IDs in group 11
group11_ids <- time_adjusted_data_for_lm %>%
  filter(sampling_phase != "Phase 3", clust_traj12 == "11") %>%
  distinct(idnr) %>% pull(idnr)

# Function for subsampling iteration
run_one_subsample <- function(i) {
  set.seed(seed_base + i)
  
  # randomly sample n12 subjects from group11 (with replacement)
  sampled_11 <- sample(group11_ids, size = n12, replace = TRUE)
  
  # build data_list for analysis: keep group12 and the sampled group11 subjects
  data_subset <- time_adjusted_data_for_lm %>%
    filter(sampling_phase != "Phase 3") %>%
    filter(clust_traj12 == "12" | idnr %in% sampled_11)
  
  data_list <- list(
    values = data_subset %>%
      dplyr::select(sample_id, all_of(olink_proteins_ta)),
    sinfo = data_subset %>%
      dplyr::select(idnr, sample_id, sampling_phase, clust_traj12, male)
  )
  
  # Run within-phase comparison:
  # lm_11_12 <- lm_analysis(data_list, "sample_id", predictor = "clust_traj12", covar = "male", stratify_by = "sampling_phase")
  
  # Run the delta-delta analysis (difference between in change between Phase1 and Phase2 for each group)
  lm_dd <- lm_analysis(
    data_list,
    "sample_id",
    predictor = "clust_traj12",
    covar = "male",
    difference_col = "sampling_phase",
    individual_id = "idnr",
    differences_between = c("Phase 1", "Phase 2")
  )
  
  # lm_dd$results expected to be a data.frame with columns: assay, estimate, pvalue, fdr
  res <- lm_dd$results %>%
    mutate(iter = i) %>%
    dplyr::select(assay, median_diff, estimate, p_value, fdr, iter)
  
  # Return list of significant assays this run
  sig_assays <- res %>% filter(fdr < fdr_cutoff) %>% pull(assay)
  
  list(results = res, significant = sig_assays)
}



# Run many iterations (serial)
pb <- progress_bar$new(total = n_runs)
all_iter_results <- vector("list", n_runs)
for (i in seq_len(n_runs)) {
  iter_out <- tryCatch({
    out <- run_one_subsample(i)
    pb$tick()
    out
  }, error = function(e) {
    pb$tick()
    message("Error in iteration ", i, ": ", e$message)
    list(results = tibble(assay = character(), median_diff = numeric(), estimate = numeric(), pvalue = numeric(), fdr = numeric(), iter = integer()),
         significant = character(0))
  })
  all_iter_results[[i]] <- iter_out
}

# -----------------------
# Aggregate outputs
# -----------------------

# 1) Frequency table: how many times each assay was significant
all_sig_lists <- map(all_iter_results, "significant")
all_sig_flat <- unlist(all_sig_lists)
protein_counts <- sort(table(all_sig_flat), decreasing = TRUE)
protein_counts_df <- tibble(
  assay = names(protein_counts),
  count = as.integer(protein_counts),
  freq = count / n_runs
)

# 2) Collect all per-iteration results (estimates, p-values)
all_results_df <-all_iter_results %>%
  map("results") %>%
  list_rbind()

# compute median estimate, proportion of runs with significant p (or fdr), direction consistency
summary_df <- all_results_df %>%
  group_by(assay) %>%
  summarise(
    median_median_diff = median( median_diff, na.rm = TRUE),
    median_estimate = median(estimate, na.rm = TRUE),
    mean_estimate = mean(estimate, na.rm = TRUE),
    prop_pos = mean(estimate > 0, na.rm = TRUE),
    prop_neg = mean(estimate < 0, na.rm = TRUE),
    prop_sig = mean(fdr < fdr_cutoff, na.rm = TRUE),
    n_non_na = sum(!is.na(estimate)),
    .groups = "drop"
  ) %>%
  arrange(desc(prop_sig), desc(prop_pos))


### Visualize it:
library(wesanderson)

# Define significance thresholds for visualization
high_confidence <- 0.8   # Proteins significant in 80%+ of runs
moderate_confidence <- 0.5 # Proteins significant in 50%+ of runs

# Create a robustness category
plot_data <- summary_df %>%
  filter(prop_sig > 0) %>%
  mutate(
    robustness = case_when(
      prop_sig >= high_confidence ~ "High confidence",
      prop_sig >= moderate_confidence ~ "Moderate confidence",
      TRUE ~ "Low confidence"
    ),
    robustness = factor(robustness, levels = c("High confidence", "Moderate confidence", "Low confidence")),
    # Create size based on prop_sig for better visual emphasis
    point_size = prop_sig * 6 + 1,
    # Label only high confidence proteins or those with very large effects
    should_label = prop_sig >= high_confidence | abs(median_estimate) > 0.4
  )


# Plot 2: Lollipop chart for top proteins (emphasizes ranking)
top_proteins <- plot_data %>%
  filter(prop_sig >= 0.01) %>%  # Only show proteins with some robustness
  slice_max(prop_sig, n = 20) %>%
  mutate(assay = reorder(assay, prop_sig))

plot2_lollipop <- ggplot(top_proteins, aes(x = prop_sig, y = assay)) +
  # Lollipop stems
  geom_segment(aes(x = 0, xend = prop_sig, y = assay, yend = assay), 
               color = "gray50", size = 0.8) +
  
  # Lollipop heads colored by direction and sized by robustness
  #geom_point(aes(color = median_estimate > 0, size = prop_sig * 6)) +
  geom_point(aes(color = median_median_diff > 0, size = prop_sig * 6)) +
  #geom_text(aes(label = sprintf("%.2f", median_estimate)), 
  #          hjust = -0.2, size = 3) +
  geom_text(aes(label = sprintf("%.2f", median_median_diff)), 
            hjust = -0.2, size = 3) +
  # Reference lines
  geom_vline(xintercept = c(moderate_confidence, high_confidence), 
             linetype = "dashed", alpha = 0.5) +
  scale_color_manual(
    values = c("TRUE" = "#D51F26", "FALSE" = "#2E86AB"),
    labels = c("Negative effect", "Positive effect"),
    name = "Direction"
  ) +
  #scale_size_identity() +
  scale_x_continuous(labels = scales::percent_format(), 
                     limits = c(0, max(top_proteins$prop_sig) * 1)) +
  labs(
    title = "Proteins with significant difference in change between subjects who gain infection vs naive",
    subtitle = "Bootstrapping naive group for 100 iterations",
    x = "Proportion of iterations significant",
    y = "Protein",
    # caption = "Numbers show median effect size"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14),
    axis.text.y = element_text(size = 9),
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  guides(size = "none")

#------------------------------------------------------------------------------------------------------------------

### Compare differences in protein levels between infected (phase 2) and non infected (phase 2) at phase 1 and phase 2:

n_runs <- 100 # 500
fdr_cutoff <- 0.05
seed_base <- 123

library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)
library(progress)

# Precompute group sizes
n12 <- time_adjusted_data_for_lm %>%
  filter(sampling_phase != "Phase 3", clust_traj12 == "12") %>%
  distinct(idnr) %>%
  nrow()

# Get random IDs in group 11
group11_ids <- time_adjusted_data_for_lm %>%
  filter(sampling_phase != "Phase 3", clust_traj12 == "11") %>%
  distinct(idnr) %>%
  pull(idnr)

# Function for bootstrap iteration
run_one_bootstrap_within_phase <- function(i) {
  set.seed(seed_base + i)
  
  # randomly sample n12 subjects from group11 (WITH replacement for bootstrap)
  sampled_11 <- sample(group11_ids, size = n12, replace = TRUE)
  
  # build data_list for analysis: keep group12 and the sampled group11 subjects
  data_subset <- time_adjusted_data_for_lm %>%
    filter(sampling_phase != "Phase 3") %>%
    filter(clust_traj12 == "12" | idnr %in% sampled_11)
  
  data_list <- list(
    values = data_subset %>%
      dplyr::select(sample_id, all_of(olink_proteins_ta)),     # adjust variable name if needed
    sinfo = data_subset %>%
      dplyr::select(idnr, sample_id, sampling_phase, clust_traj12, male)
  )
  
  # Run the delta-delta analysis (difference in change between Phase1 and Phase2 for each group)
  lm_dd <- lm_analysis(
    data_list,
    "sample_id",
    predictor = "clust_traj12",
    covar = "male",
    stratify_by = "sampling_phase"
    # difference_col = "sampling_phase",
    #  individual_id = "idnr",
    # differences_between = c("Phase 1", "Phase 2")
  )
  
  # lm_dd$results expected to be a data.frame with columns: assay, estimate, p_value, fdr
  res <- lm_dd$results %>%
    mutate(iter = i) %>%
    dplyr::select(assay, sampling_phase, median_diff, estimate, p_value, fdr, iter)
  
  # Return list of significant assays this run
  sig_assays_p1 <- res %>%filter(sampling_phase == "Phase 1") %>% filter(fdr < fdr_cutoff) %>% pull(assay)
  
  sig_assays_p2 <- res %>%filter(sampling_phase == "Phase 2") %>% filter(fdr < fdr_cutoff) %>% pull(assay)
  
  list(results = res, significant_p1 = sig_assays_p1, significant_p2 = sig_assays_p2)
}

# Run many iterations (serial)
pb <- progress_bar$new(total = n_runs)
all_iter_results <- vector("list", n_runs)
for (i in seq_len(n_runs)) {
  iter_out <- tryCatch({
    out <- run_one_bootstrap_within_phase(i)
    pb$tick()
    out
  }, error = function(e) {
    pb$tick()
    message("Error in iteration ", i, ": ", e$message)
    list(results = tibble(assay = character(), median_diff = numeric(), estimate = numeric(), p_value = numeric(), fdr = numeric(), iter = integer()),
         significant_p1 = character(0),
         significant_p2 = character(0))
  })
  all_iter_results[[i]] <- iter_out
}

# Aggregate bootstrap outputs
# Frequency table: how many times each assay was significant
all_sig_lists_p1 <- map(all_iter_results, "significant_p1")
all_sig_flat_p1 <- unlist(all_sig_lists_p1)
protein_counts <- sort(table(all_sig_flat_p1), decreasing = TRUE)
protein_counts_p1_df <- tibble(
  assay = names(protein_counts),
  count = as.integer(protein_counts),
  freq = count / n_runs
)

all_sig_lists_p2 <- map(all_iter_results, "significant_p2")
all_sig_flat_p2 <- unlist(all_sig_lists_p2)
protein_counts <- sort(table(all_sig_flat_p2), decreasing = TRUE)
protein_counts_p2_df <- tibble(
  assay = names(protein_counts),
  count = as.integer(protein_counts),
  freq = count / n_runs
)


# Collect all per-iteration results (estimates, p-values)
all_results_df <- all_iter_results %>%
  map("results") %>%
  list_rbind()

# Compute summary statistics and 95% confidence intervals per assay
summary_df <- all_results_df %>%
  group_by(sampling_phase, assay) %>%
  summarise(
    median_median_diff = median( median_diff, na.rm = TRUE),
    median_estimate = median(estimate, na.rm = TRUE),
    mean_estimate = mean(estimate, na.rm = TRUE),
    lower_95 = quantile(estimate, 0.025, na.rm = TRUE),
    upper_95 = quantile(estimate, 0.975, na.rm = TRUE),
    prop_pos = mean(estimate > 0, na.rm = TRUE),
    prop_neg = mean(estimate < 0, na.rm = TRUE),
    prop_sig = mean(fdr < fdr_cutoff, na.rm = TRUE),
    n_non_na = sum(!is.na(estimate)),
    fdr_median = median(fdr),
    fdr_mean = mean(fdr),
    .groups = "drop"
  ) %>%
  arrange(desc(prop_sig), desc(prop_pos))

summary_df_1112 <- summary_df
### Plot Volcano plot:
library(ggh4x)
volcano_1112_baseline <- ggplot(summary_df_1112, aes( x = median_median_diff, y = -log10(fdr_median))) +
  geom_point( aes( color = prop_sig, size = prop_sig)) +
  facet_wrap(~sampling_phase) +
  theme_classic() +
  #theme_bw() +
  geom_text_repel(
    data = subset(summary_df_1112, prop_sig > 0.1),
    aes(label = str_remove(assay, "_Phase \\d")),  # Use the `protein` column as the label
    color = "black",
    box.padding = 0.5,
    max.overlaps = Inf
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  #colr_scale +
  scale_x_continuous(limits = center_limits()) +
  scale_color_gradientn(
    colors = c("grey80", wes_palette("Zissou1", 100, type = "continuous")),
    values = scales::rescale(c(0, 0.0001, 1)),  # Grey for 0, then gradient starts
    limits = c(0, 1),
    name = "Proportion\nsignificant"
  ) +
  scale_y_continuous(breaks = round(seq(0, max(-log10(summary_df$fdr_median)+2), by = 1),1)) +
  labs(
    title = "Subjects who gain infection in phase 2 vs naive",
    subtitle = "Bootstrapping naive group for 100 iterations",
    x = "Median difference between groups",
    y = "-log10(FDR)"
  ) +
  geom_vline( xintercept = 0, linetype = "dashed", color = "grey")




#----------------------------------------------------------------------------------------------------------------
### Plot protein level and calculate significant using wilcox:
## Calculate statistical result function:

# Function to perform statistical tests for each protein
calculate_protein_stats <- # function(proteins_of_interest, data) {
  function(proteins_of_interest_tad, data) {
    
    results_list <- list()
    
    #for(protein in proteins_of_interest) {
    for(protein in proteins_of_interest_tad) {
      
      cat("Processing protein:", protein, "\n")
      
      # Prepare data for this protein
      protein_data <- data %>% 
        filter(idnr %in% sminet_infected_subjects) %>%
        mutate(
          time_since_sminet = as.Date(sampling_time) - as.Date(sminet_date),
          time_since_sminet_group = case_when(
            time_since_sminet < 0 ~ "pre",
            time_since_sminet >= 0 & time_since_sminet <= 30 ~ "0-30",
            time_since_sminet > 30 ~ "30+"
          ),
          time_since_sminet_group = factor(time_since_sminet_group, levels = c("pre", "0-30", "30+"))
        ) %>%
        filter(Time_Sample_Dose1 < 0) %>%
        filter( ! sampling_phase == "Phase 3") |> # this is only one sample. remove so we only have one pre and post sample per subject
        pivot_longer(cols = all_of(proteins_of_interest[-grep("PKLR|GLOD4|DNPH1", proteins_of_interest)]), names_to = "assay", values_to = "npx") %>%
        filter(assay == protein)
      
      # Initialize results for this protein
      protein_results <- data.frame(
        protein = protein,
        p_value_pre_vs_0_30 = NA,
        eff_size_pre_vs_0_30 = NA,     # <-- Added
        p_value_pre_vs_30plus = NA,
        eff_size_pre_vs_30plus = NA,   # <-- Added
        p_value_0_30_vs_30plus = NA,
        eff_size_0_30_vs_30plus = NA,  # <-- Added
        n_pre = 0,
        n_0_30 = 0,
        n_30plus = 0,
        n_paired_pre_0_30 = 0,
        n_paired_pre_30plus = 0,
        test_type_pre_0_30 = NA,
        test_type_pre_30plus = NA,
        test_type_0_30_30plus = NA
      )
      
      # Count samples in each group
      protein_results$n_pre <- sum(protein_data$time_since_sminet_group == "pre", na.rm = TRUE)
      protein_results$n_0_30 <- sum(protein_data$time_since_sminet_group == "0-30", na.rm = TRUE)
      protein_results$n_30plus <- sum(protein_data$time_since_sminet_group == "30+", na.rm = TRUE)
      
      # Test 1: pre vs 0-30 (paired if same subjects exist in both groups)
      pre_data <- protein_data %>% filter(time_since_sminet_group == "pre")
      group_0_30_data <- protein_data %>% filter(time_since_sminet_group == "0-30")
      
      # Find subjects present in both pre and 0-30 groups
      paired_subjects_pre_0_30 <- intersect(pre_data$idnr, group_0_30_data$idnr)
      protein_results$n_paired_pre_0_30 <- length(paired_subjects_pre_0_30)
      
      if(length(paired_subjects_pre_0_30) >= 5) {  # Minimum 5 pairs for meaningful paired test
        # Paired Wilcoxon
        pre_paired <- pre_data %>% filter(idnr %in% paired_subjects_pre_0_30) %>% arrange(idnr)
        group_0_30_paired <- group_0_30_data %>% filter(idnr %in% paired_subjects_pre_0_30) %>% arrange(idnr)
        
        if(nrow(pre_paired) == nrow(group_0_30_paired)) {
          test_result <- wilcox.test(x = pre_paired$npx, y = group_0_30_paired$npx, paired = TRUE, alternative = "two.sided")
          protein_results$p_value_pre_vs_0_30 <- test_result$p.value
          protein_results$eff_size_pre_vs_0_30 <- get_wilcox_effect_size(pre_paired$npx, group_0_30_paired$npx, paired = TRUE) #new
          protein_results$test_type_pre_0_30 = "paired_wilcoxon"
        }
      } else if(nrow(pre_data) >= 3 && nrow(group_0_30_data) >= 3) {
        # Unpaired t-test if not enough pairs
        # test_result <- t.test(pre_data$npx, group_0_30_data$npx, paired = FALSE)
        test_result <- wilcox.test(pre_data$npx, group_0_30_data$npx, paired = FALSE)
        protein_results$p_value_pre_vs_0_30 <- test_result$p.value
        protein_results$eff_size_pre_vs_0_30 <- get_wilcox_effect_size(pre_data$npx, group_0_30_data$npx, paired = FALSE) #new
        protein_results$test_type_pre_0_30 = "unpaired_wilcoxon"
      }
      
      # Test 2: pre vs 30+ (paired if same subjects exist in both groups)
      group_30plus_data <- protein_data %>% filter(time_since_sminet_group == "30+")
      
      # Find subjects present in both pre and 30+ groups
      paired_subjects_pre_30plus <- intersect(pre_data$idnr, group_30plus_data$idnr)
      protein_results$n_paired_pre_30plus <- length(paired_subjects_pre_30plus)
      
      if(length(paired_subjects_pre_30plus) >= 5) {  # Minimum 5 pairs for meaningful paired test
        # Paired Wilcoxon
        pre_paired <- pre_data %>% filter(idnr %in% paired_subjects_pre_30plus) %>% arrange(idnr)
        group_30plus_paired <- group_30plus_data %>% filter(idnr %in% paired_subjects_pre_30plus) %>% arrange(idnr)
        
        if(nrow(pre_paired) == nrow(group_30plus_paired)) {
          test_result <- wilcox.test(pre_paired$npx, group_30plus_paired$npx, paired = TRUE )
          protein_results$p_value_pre_vs_30plus <- test_result$p.value
          protein_results$eff_size_pre_vs_30plus <- get_wilcox_effect_size(pre_paired$npx, group_30plus_paired$npx, paired = TRUE) #new
          protein_results$test_type_pre_30plus = "paired_wilcoxon"
        }
      } else if(nrow(pre_data) >= 3 && nrow(group_30plus_data) >= 3) {
        test_result <- wilcox.test(pre_data$npx, group_30plus_data$npx, paired = FALSE)
        protein_results$p_value_pre_vs_30plus <- test_result$p.value
        protein_results$eff_size_pre_vs_30plus <- get_wilcox_effect_size(pre_data$npx, group_30plus_data$npx, paired = FALSE)
        protein_results$test_type_pre_30plus = "unpaired_wilcoxon"
      }
      
      # Test 3: 0-30 vs 30+ (unpaired - different subjects)
      if(nrow(group_0_30_data) >= 3 && nrow(group_30plus_data) >= 3) {
        test_result <- wilcox.test(group_0_30_data$npx, group_30plus_data$npx, paired = FALSE)
        protein_results$p_value_0_30_vs_30plus <- test_result$p.value
        protein_results$eff_size_0_30_vs_30plus <- get_wilcox_effect_size(group_0_30_data$npx, group_30plus_data$npx, paired = FALSE) #new
        protein_results$test_type_0_30_30plus = "unpaired_wilcoxon"
      }
      
      results_list[[protein]] <- protein_results
    }
    
    # Combine results
    all_results <- do.call(rbind, results_list)
    
    # Apply FDR correction to each p-value column
    all_results$p_adj_pre_vs_0_30 <- p.adjust(all_results$p_value_pre_vs_0_30, method = "fdr")
    all_results$p_adj_pre_vs_30plus <- p.adjust(all_results$p_value_pre_vs_30plus, method = "fdr")
    all_results$p_adj_0_30_vs_30plus <- p.adjust(all_results$p_value_0_30_vs_30plus, method = "fdr")
    
    # Add significance indicators
    all_results$sig_pre_vs_0_30 <- case_when(
      all_results$p_adj_pre_vs_0_30 < 0.001 ~ "***",
      all_results$p_adj_pre_vs_0_30 < 0.01 ~ "**",
      all_results$p_adj_pre_vs_0_30 < 0.05 ~ "*",
      TRUE ~ "ns"
    )
    
    all_results$sig_pre_vs_30plus <- case_when(
      all_results$p_adj_pre_vs_30plus < 0.001 ~ "***",
      all_results$p_adj_pre_vs_30plus < 0.01 ~ "**",
      all_results$p_adj_pre_vs_30plus < 0.05 ~ "*",
      TRUE ~ "ns"
    )
    
    all_results$sig_0_30_vs_30plus <- case_when(
      all_results$p_adj_0_30_vs_30plus < 0.001 ~ "***",
      all_results$p_adj_0_30_vs_30plus < 0.01 ~ "**",
      all_results$p_adj_0_30_vs_30plus < 0.05 ~ "*",
      TRUE ~ "ns"
    )
    
    return(all_results)
    
  } ## END function

##                             Plot function for pre and post infection exl vaccinated:
create_plots_with_ggsignif <- function(proteins_of_interest, data, statistical_results) {
  
  # You'll need: install.packages("ggsignif")
  library(ggsignif)
  
  plot_list <- lapply(proteins_of_interest, function(protein) {
    
    # Get statistical results for this protein
    protein_stats <- statistical_results %>% 
      filter(protein == !!protein)
    
    # Prepare plot data
    plot_data <- data %>% 
      filter(idnr %in% sminet_infected_subjects) %>%
      mutate(
        time_since_sminet = as.Date(sampling_time) - as.Date(sminet_date),
        time_since_sminet_group = case_when(
          time_since_sminet < 0 ~ "pre",
          time_since_sminet >= 0 & time_since_sminet <= 30 ~ "0-30",
          time_since_sminet > 30 ~ "30+"
        ),
        time_since_sminet_group = factor(time_since_sminet_group, levels = c("pre", "0-30",  "30+"))
      ) %>%
      filter(Time_Sample_Dose1 < 0) %>%
      filter( ! sampling_phase == "Phase 3") |>
      pivot_longer(cols = all_of(proteins_of_interest), names_to = "assay", values_to = "npx") %>%
      #pivot_longer(cols = proteins_of_interest, names_to = "assay", values_to = "npx") %>%
      filter(assay == protein)
    
    # Prepare significance annotations with test info
    comparisons <- list()
    annotations <- c()
    
    if(!is.na(protein_stats$p_adj_pre_vs_0_30)) {
      comparisons <- append(comparisons, list(c("pre", "0-30")))
      annotations <- c(annotations, paste0(
        protein_stats$sig_pre_vs_0_30, "\n(", protein_stats$test_type_pre_0_30, ")"
      ))
    }
    
    if(!is.na(protein_stats$p_adj_pre_vs_30plus)) {
      comparisons <- append(comparisons, list(c("pre", "30+")))
      annotations <- c(annotations, paste0(
        protein_stats$sig_pre_vs_30plus, "\n(", protein_stats$test_type_pre_30plus, ")"
      ))
    }
    
    if(!is.na(protein_stats$p_adj_0_30_vs_30plus)) {
      comparisons <- append(comparisons, list(c("0-30", "30+")))
      annotations <- c(annotations, paste0(
        protein_stats$sig_0_30_vs_30plus, "\n(", protein_stats$test_type_0_30_30plus, ")"
      ))
    }
  
    p1 <- ggplot(plot_data, aes(x = time_since_sminet_group, y = npx, group = idnr)) +
      geom_beeswarm(aes( my_col = time_since_sminet_group, my_x = time_since_sminet_group))
    pb <- ggplot_build(p1)
    
    p <- ggplot(pb$data[[1]], aes(x = x, y = y, group = group)) +
      geom_line(aes(group = group), col = "grey80", alpha = 0.2) +
      geom_point( aes( color = my_col), size = 3, alpha = 0.6) +
      theme_classic() +
      scale_x_discrete() +
      labs(title = protein, y = "ProtPQN NPX [AU]", x = "Days since PCR+", subtitle = "Subjects with PCR+ confirmed infection, no vaccination") +
      geom_boxplot(aes(x = my_x, y = y, group = my_x), outlier.shape = NA, fill = NA) +
      scale_color_manual( values = c("grey", "red", "orange")) +
      guides(color="none")
    
    
    # Add significance brackets if we have comparisons
    if(length(comparisons) > 0) {
      p <- p + geom_signif(
        comparisons = comparisons,
        annotations = annotations,
        step_increase = 0.1,
        textsize = 2.5
      )
    }
    
    return(p)
  })
  
  names(plot_list) <- proteins_of_interest
  return(plot_list)
}


## Call function to create the plots:
proteins_of_interest_tad <- proteins_of_interest[-grep("PKLR|GLOD4|DNPH1", proteins_of_interest )]

statistical_results_timeadjust <- calculate_protein_stats(proteins_of_interest_tad, time_adjusted_data_si)
enhanced_plots_timeadjust <- create_plots_with_ggsignif(proteins_of_interest_tad, time_adjusted_data_si, statistical_results_timeadjust)

########## Protein levels pre and post infection but include vaccinated samples:
# Function to perform statistical tests for each protein
calculate_protein_stats_vi <- # function(proteins_of_interest, data) {
  function(proteins_of_interest_tad, data) {
    
    results_list <- list()
    
    #for(protein in proteins_of_interest) {
    for(protein in proteins_of_interest_tad) {
      
      cat("Processing protein:", protein, "\n")
      
      # Prepare data for this protein
      protein_data <- data %>% 
        filter(idnr %in% sminet_infected_subjects) %>%
        mutate(
          time_since_sminet = as.Date(sampling_time) - as.Date(sminet_date),
          time_since_sminet_group = case_when(
            time_since_sminet < 0 ~ "pre",
            time_since_sminet >= 0 & time_since_sminet <= 30 ~ "0-30",
            time_since_sminet > 30 ~ "30+"
          ),
          time_since_sminet_group = factor(time_since_sminet_group, levels = c("pre", "0-30", "30+"))
        ) %>%
        pivot_longer(cols = all_of(proteins_of_interest), names_to = "assay", values_to = "npx") %>%
        filter(assay == protein)
      
      # Initialize results for this protein
      protein_results <- data.frame(
        protein = protein,
        p_value_pre_vs_0_30 = NA,
        eff_size_pre_vs_0_30 = NA,     # <-- Added
        p_value_pre_vs_30plus = NA,
        eff_size_pre_vs_30plus = NA,   # <-- Added
        p_value_0_30_vs_30plus = NA,
        eff_size_0_30_vs_30plus = NA,  # <-- Added
        n_pre = 0,
        n_0_30 = 0,
        n_30plus = 0,
        n_paired_pre_0_30 = 0,
        n_paired_pre_30plus = 0,
        test_type_pre_0_30 = NA,
        test_type_pre_30plus = NA,
        test_type_0_30_30plus = NA
      )
      
      # Count samples in each group
      protein_results$n_pre <- sum(protein_data$time_since_sminet_group == "pre", na.rm = TRUE)
      protein_results$n_0_30 <- sum(protein_data$time_since_sminet_group == "0-30", na.rm = TRUE)
      protein_results$n_30plus <- sum(protein_data$time_since_sminet_group == "30+", na.rm = TRUE)
      
      # Test 1: pre vs 0-30 (paired if same subjects exist in both groups)
      pre_data <- protein_data %>% filter(time_since_sminet_group == "pre")
      group_0_30_data <- protein_data %>% filter(time_since_sminet_group == "0-30")
      
      # Find subjects present in both pre and 0-30 groups
      paired_subjects_pre_0_30 <- intersect(pre_data$idnr, group_0_30_data$idnr)
      protein_results$n_paired_pre_0_30 <- length(paired_subjects_pre_0_30)
      test_result <- wilcox.test(pre_data$npx, group_0_30_data$npx, paired = FALSE)
      protein_results$p_value_pre_vs_0_30 <- test_result$p.value
      protein_results$eff_size_pre_vs_0_30 <- get_wilcox_effect_size(pre_data$npx, group_0_30_data$npx, paired = FALSE) #new
      protein_results$test_type_pre_0_30 = "unpaired_wilcoxon"
      
      # }
      
      # Test 2: pre vs 30+ (paired if same subjects exist in both groups)
      group_30plus_data <- protein_data %>% filter(time_since_sminet_group == "30+")
      
      # Find subjects present in both pre and 30+ groups
      paired_subjects_pre_30plus <- intersect(pre_data$idnr, group_30plus_data$idnr)
      protein_results$n_paired_pre_30plus <- length(paired_subjects_pre_30plus)
      
      #if(length(paired_subjects_pre_30plus) >= 5) {  # Minimum 5 pairs for meaningful paired test
      # Paired t-test
      pre_paired <- pre_data %>% filter(idnr %in% paired_subjects_pre_30plus) %>% arrange(idnr)
      group_30plus_paired <- group_30plus_data %>% filter(idnr %in% paired_subjects_pre_30plus) %>% arrange(idnr)
      
      test_result <- wilcox.test(pre_data$npx, group_30plus_data$npx, paired = FALSE)
      protein_results$p_value_pre_vs_30plus <- test_result$p.value
      protein_results$eff_size_pre_vs_30plus <- get_wilcox_effect_size(pre_data$npx, group_30plus_data$npx, paired = FALSE)
      protein_results$test_type_pre_30plus = "unpaired_wilcoxon"
      #  }
      
      # Test 3: 0-30 vs 30+ (unpaired - different subjects)
      if(nrow(group_0_30_data) >= 3 && nrow(group_30plus_data) >= 3) {
        #test_result <- t.test(group_0_30_data$npx, group_30plus_data$npx, paired = FALSE)
        test_result <- wilcox.test(group_0_30_data$npx, group_30plus_data$npx, paired = FALSE)
        protein_results$p_value_0_30_vs_30plus <- test_result$p.value
        protein_results$eff_size_0_30_vs_30plus <- get_wilcox_effect_size(group_0_30_data$npx, group_30plus_data$npx, paired = FALSE) #new
        protein_results$test_type_0_30_30plus = "unpaired_wilcoxon"
      }
      
      results_list[[protein]] <- protein_results
    }
    
    # Combine results
    all_results <- do.call(rbind, results_list)
    
    # Apply FDR correction to each p-value column
    all_results$p_adj_pre_vs_0_30 <- p.adjust(all_results$p_value_pre_vs_0_30, method = "fdr")
    all_results$p_adj_pre_vs_30plus <- p.adjust(all_results$p_value_pre_vs_30plus, method = "fdr")
    all_results$p_adj_0_30_vs_30plus <- p.adjust(all_results$p_value_0_30_vs_30plus, method = "fdr")
    
    # Add significance indicators
    all_results$sig_pre_vs_0_30 <- case_when(
      all_results$p_adj_pre_vs_0_30 < 0.001 ~ "***",
      all_results$p_adj_pre_vs_0_30 < 0.01 ~ "**",
      all_results$p_adj_pre_vs_0_30 < 0.05 ~ "*",
      TRUE ~ "ns"
    )
    
    all_results$sig_pre_vs_30plus <- case_when(
      all_results$p_adj_pre_vs_30plus < 0.001 ~ "***",
      all_results$p_adj_pre_vs_30plus < 0.01 ~ "**",
      all_results$p_adj_pre_vs_30plus < 0.05 ~ "*",
      TRUE ~ "ns"
    )
    
    all_results$sig_0_30_vs_30plus <- case_when(
      all_results$p_adj_0_30_vs_30plus < 0.001 ~ "***",
      all_results$p_adj_0_30_vs_30plus < 0.01 ~ "**",
      all_results$p_adj_0_30_vs_30plus < 0.05 ~ "*",
      TRUE ~ "ns"
    )
    
    return(all_results)
  }


# Plot
create_plots_with_ggsignif_vi <- function(proteins_of_interest, data, statistical_results) {
  
  # You'll need: install.packages("ggsignif")
  library(ggsignif)
  
  plot_list <- lapply(proteins_of_interest, function(protein) {
    
    # Get statistical results for this protein
    protein_stats <- statistical_results %>% 
      filter(protein == !!protein)
    
    # Prepare plot data
    plot_data <- data %>% 
      filter(idnr %in% sminet_infected_subjects) %>%
      mutate(
        time_since_sminet = as.Date(sampling_time) - as.Date(sminet_date),
        time_since_sminet_group = case_when(
          time_since_sminet < 0 ~ "pre",
          time_since_sminet >= 0 & time_since_sminet <= 30 ~ "0-30",
          time_since_sminet > 30 ~ "30+"),
        vaccinated = case_when( Time_Sample_Dose1 <= 0 ~ "No",
                                T ~ "Yes"),
        time_since_sminet_group = factor(time_since_sminet_group, levels = c("pre", "0-30",  "30+"))
      ) %>%
      #filter(Time_Sample_Dose1 < 0) %>%
      #     filter( ! sampling_phase == "Phase 3") |>
      pivot_longer(cols = all_of(proteins_of_interest), names_to = "assay", values_to = "npx") %>%
      #pivot_longer(cols = proteins_of_interest, names_to = "assay", values_to = "npx") %>%
      filter(assay == protein)
    
    # Prepare significance annotations with test info
    comparisons <- list()
    annotations <- c()
    
    if(!is.na(protein_stats$p_adj_pre_vs_0_30)) {
      comparisons <- append(comparisons, list(c("pre", "0-30")))
      annotations <- c(annotations, paste0(
        protein_stats$sig_pre_vs_0_30, "\n(", protein_stats$test_type_pre_0_30, ")"
      ))
    }
    
    if(!is.na(protein_stats$p_adj_pre_vs_30plus)) {
      comparisons <- append(comparisons, list(c("pre", "30+")))
      annotations <- c(annotations, paste0(
        protein_stats$sig_pre_vs_30plus, "\n(", protein_stats$test_type_pre_30plus, ")"
      ))
    }
    
    if(!is.na(protein_stats$p_adj_0_30_vs_30plus)) {
      comparisons <- append(comparisons, list(c("0-30", "30+")))
      annotations <- c(annotations, paste0(
        protein_stats$sig_0_30_vs_30plus, "\n(", protein_stats$test_type_0_30_30plus, ")"
      ))
    }
    
    p1 <- ggplot(plot_data, aes(x = time_since_sminet_group, y = npx, group = idnr)) +
      geom_beeswarm(aes( my_col = time_since_sminet_group, my_x = time_since_sminet_group, my_shape = vaccinated))
    pb <- ggplot_build(p1)
    
    p <- ggplot(pb$data[[1]], aes(x = x, y = y, group = group)) +
      geom_line(aes(group = group), col = "grey80", alpha = 0.2) +
      geom_point( aes( color = my_col, shape = my_shape), size = 3, alpha = 0.6) +
      theme_classic() +
      scale_x_discrete() +
      labs(title = protein, y = "ProtPQN NPX [AU]", x = "Days since PCR+", subtitle = "Subjects with PCR+ confirmed infection and possibly vaccinated subjects") +
      geom_boxplot(aes(x = my_x, y = y, group = my_x), outlier.shape = NA, fill = NA) +
      scale_color_manual( values = c("grey", "red", "orange")) +
      guides(color="none")
    
    # Add significance brackets if we have comparisons
    if(length(comparisons) > 0) {
      p <- p + geom_signif(
        comparisons = comparisons,
        annotations = annotations,
        step_increase = 0.1,
        textsize = 2.5
      )
    }
    
    return(p)
  })
  
  names(plot_list) <- proteins_of_interest
  return(plot_list)
}

# Create the plots:
statistical_results_timeadjust_vi <- calculate_protein_stats_vi(proteins_of_interest_tad, time_adjusted_data_si)
enhanced_plots_timeadjust_vi <- create_plots_with_ggsignif_vi(proteins_of_interest_tad, time_adjusted_data_si, statistical_results_timeadjust_vi)

#----------------------------------------------------------------------------------------------------------------
### Plot protein level for vaccinated and calculate significant using wilcox:
calculate_protein_stats_vaccine <- function(proteins_of_interest_tad, data) {
    
    results_list <- list()

    for(protein in proteins_of_interest_tad) {
      
      cat("Processing protein:", protein, "\n")
      
      # Prepare data for this protein
      protein_data <- data %>% 
        filter(!cluster_maximum_6prot_omi %in% c("2", "3")) |>
        mutate(
          time_since_vaccination_group = case_when(
            time_since_closest_date < 0 ~ "pre",
            time_since_closest_date >= 0 & time_since_closest_date <= 30 ~ "0-30",
            time_since_closest_date > 30 ~ "30+"
          ),
          time_since_vaccination_group = factor(time_since_vaccination_group, levels = c("pre", "0-30", "30+"))
        ) %>%
        pivot_longer(cols = all_of(proteins_of_interest), names_to = "assay", values_to = "npx") %>%
        filter(assay == protein)
      
      # Initialize results for this protein
      protein_results <- data.frame(
        protein = protein,
        p_value_pre_vs_0_30 = NA,
        eff_size_pre_vs_0_30 = NA,     # <-- Added
        p_value_pre_vs_30plus = NA,
        eff_size_pre_vs_30plus = NA,   # <-- Added
        p_value_0_30_vs_30plus = NA,
        eff_size_0_30_vs_30plus = NA,  # <-- Added
        n_pre = 0,
        n_0_30 = 0,
        n_30plus = 0,
        n_paired_pre_0_30 = 0,
        n_paired_pre_30plus = 0,
        test_type_pre_0_30 = NA,
        test_type_pre_30plus = NA,
        test_type_0_30_30plus = NA
      )
      
      # Count samples in each group
      protein_results$n_pre <- sum(protein_data$time_since_vaccination_group == "pre", na.rm = TRUE)
      protein_results$n_0_30 <- sum(protein_data$time_since_vaccination_group == "0-30", na.rm = TRUE)
      protein_results$n_30plus <- sum(protein_data$time_since_vaccination_group == "30+", na.rm = TRUE)
      
      # Test 1: pre vs 0-30 (paired if same subjects exist in both groups)
      pre_data <- protein_data %>% filter(time_since_vaccination_group == "pre")
      group_0_30_data <- protein_data %>% filter(time_since_vaccination_group == "0-30")
      
      # Find subjects present in both pre and 0-30 groups
      paired_subjects_pre_0_30 <- intersect(pre_data$idnr, group_0_30_data$idnr)
      protein_results$n_paired_pre_0_30 <- length(paired_subjects_pre_0_30)
      
      if(length(paired_subjects_pre_0_30) >= 5) {  # Minimum 5 pairs for meaningful paired test
        # Paired t-test
        pre_paired <- pre_data %>% filter(idnr %in% paired_subjects_pre_0_30) %>% arrange(idnr)
        group_0_30_paired <- group_0_30_data %>% filter(idnr %in% paired_subjects_pre_0_30) %>% arrange(idnr)
        
        if(nrow(pre_paired) == nrow(group_0_30_paired)) {
          #test_result <- t.test(pre_paired$npx, group_0_30_paired$npx, paired = TRUE)
          test_result <- wilcox.test(pre_paired$npx, group_0_30_paired$npx, paired = TRUE)
          protein_results$p_value_pre_vs_0_30 <- test_result$p.value
          protein_results$eff_size_pre_vs_0_30 <- get_wilcox_effect_size(pre_paired$npx, group_0_30_paired$npx, paired = TRUE) #new
          protein_results$test_type_pre_0_30 = "paired_wilcoxon"
        }
      } else if(nrow(pre_data) >= 3 && nrow(group_0_30_data) >= 3) {
        # Unpaired t-test if not enough pairs
        #test_result <- t.test(pre_data$npx, group_0_30_data$npx, paired = FALSE)
        test_result <- wilcox.test(pre_data$npx, group_0_30_data$npx, paired = FALSE)
        protein_results$p_value_pre_vs_0_30 <- test_result$p.value
        protein_results$eff_size_pre_vs_0_30 <- get_wilcox_effect_size(pre_data$npx, group_0_30_data$npx, paired = FALSE) #new
        protein_results$test_type_pre_0_30 = "unpaired_wilcoxon"
      }
      
      # Test 2: pre vs 30+ (paired if same subjects exist in both groups)
      group_30plus_data <- protein_data %>% filter(time_since_vaccination_group == "30+")
      
      # Find subjects present in both pre and 30+ groups
      paired_subjects_pre_30plus <- intersect(pre_data$idnr, group_30plus_data$idnr)
      protein_results$n_paired_pre_30plus <- length(paired_subjects_pre_30plus)
      
      if(length(paired_subjects_pre_30plus) >= 5) {  # Minimum 5 pairs for meaningful paired test
        # Paired t-test
        pre_paired <- pre_data %>% filter(idnr %in% paired_subjects_pre_30plus) %>% arrange(idnr)
        group_30plus_paired <- group_30plus_data %>% filter(idnr %in% paired_subjects_pre_30plus) %>% arrange(idnr)
        
        if(nrow(pre_paired) == nrow(group_30plus_paired)) {
          test_result <- wilcox.test(pre_paired$npx, group_30plus_paired$npx, paired = TRUE)
          protein_results$p_value_pre_vs_30plus <- test_result$p.value
          protein_results$eff_size_pre_vs_30plus <- get_wilcox_effect_size(pre_paired$npx, group_30plus_paired$npx, paired = TRUE) #new
          protein_results$test_type_pre_30plus = "paired_wilcoxon"
          
        } else { #(nrow(pre_data) >= 3 && nrow(group_30plus_data) >= 3) {
          test_result <- wilcox.test(pre_data$npx, group_30plus_data$npx, paired = FALSE)
          protein_results$p_value_pre_vs_30plus <- test_result$p.value
          protein_results$eff_size_pre_vs_30plus <- get_wilcox_effect_size(pre_data$npx, group_30plus_data$npx, paired = FALSE)
          protein_results$test_type_pre_30plus = "unpaired_wilcoxon"
          
          
        }
      }
      
      # Test 3: 0-30 vs 30+ (unpaired - different subjects)
      if(nrow(group_0_30_data) >= 3 && nrow(group_30plus_data) >= 3) {
        #test_result <- t.test(group_0_30_data$npx, group_30plus_data$npx, paired = FALSE)
        test_result <- wilcox.test(group_0_30_data$npx, group_30plus_data$npx, paired = FALSE)
        protein_results$p_value_0_30_vs_30plus <- test_result$p.value
        protein_results$eff_size_0_30_vs_30plus <- get_wilcox_effect_size(group_0_30_data$npx, group_30plus_data$npx, paired = FALSE) #new
        protein_results$test_type_0_30_30plus = "unpaired_wilcoxon"
      }
      
      results_list[[protein]] <- protein_results
    }
    
    # Combine results
    all_results <- do.call(rbind, results_list)
    
    # Apply FDR correction to each p-value column
    all_results$p_adj_pre_vs_0_30 <- p.adjust(all_results$p_value_pre_vs_0_30, method = "fdr")
    all_results$p_adj_pre_vs_30plus <- p.adjust(all_results$p_value_pre_vs_30plus, method = "fdr")
    all_results$p_adj_0_30_vs_30plus <- p.adjust(all_results$p_value_0_30_vs_30plus, method = "fdr")
    
    # Add significance indicators
    all_results$sig_pre_vs_0_30 <- case_when(
      all_results$p_adj_pre_vs_0_30 < 0.001 ~ "***",
      all_results$p_adj_pre_vs_0_30 < 0.01 ~ "**",
      all_results$p_adj_pre_vs_0_30 < 0.05 ~ "*",
      TRUE ~ "ns"
    )
    
    all_results$sig_pre_vs_30plus <- case_when(
      all_results$p_adj_pre_vs_30plus < 0.001 ~ "***",
      all_results$p_adj_pre_vs_30plus < 0.01 ~ "**",
      all_results$p_adj_pre_vs_30plus < 0.05 ~ "*",
      TRUE ~ "ns"
    )
    
    all_results$sig_0_30_vs_30plus <- case_when(
      all_results$p_adj_0_30_vs_30plus < 0.001 ~ "***",
      all_results$p_adj_0_30_vs_30plus < 0.01 ~ "**",
      all_results$p_adj_0_30_vs_30plus < 0.05 ~ "*",
      TRUE ~ "ns"
    )
    
    return(all_results)
  }
# 


create_plots_with_ggsignif_vaccine <- function(proteins_of_interest, data, statistical_results) {
  
  # You'll need: install.packages("ggsignif")
  library(ggsignif)
  
  plot_list <- lapply(proteins_of_interest, function(protein) {
    
    # Get statistical results for this protein
    protein_stats <- statistical_results %>% 
      filter(protein == !!protein)
    
    # Prepare plot data
    plot_data <- data %>% 
      filter(!cluster_maximum_6prot_omi %in% c("2", "3")) |>
      mutate(
        # time_since_sminet = as.Date(sampling_time) - as.Date(sminet_date),
        time_since_vaccination_group = case_when(
          time_since_closest_date < 0 ~ "pre",
          time_since_closest_date >= 0 & time_since_closest_date <= 30 ~ "0-30",
          time_since_closest_date > 30 ~ "30+"
        ),
        time_since_vaccination_group = factor(time_since_vaccination_group, levels = c("pre", "0-30", "30+"))
      ) %>%
      pivot_longer(cols = all_of(proteins_of_interest), names_to = "assay", values_to = "npx") %>%
      filter(assay == protein)
    
    # Prepare significance annotations with test info
    comparisons <- list()
    annotations <- c()
    
    if(!is.na(protein_stats$p_adj_pre_vs_0_30)) {
      comparisons <- append(comparisons, list(c("pre", "0-30")))
      annotations <- c(annotations, paste0(
        protein_stats$sig_pre_vs_0_30, "\n(", protein_stats$test_type_pre_0_30, ")"
      ))
    }
    
    if(!is.na(protein_stats$p_adj_pre_vs_30plus)) {
      comparisons <- append(comparisons, list(c("pre", "30+")))
      annotations <- c(annotations, paste0(
        protein_stats$sig_pre_vs_30plus, "\n(", protein_stats$test_type_pre_30plus, ")"
      ))
    }
    
    if(!is.na(protein_stats$p_adj_0_30_vs_30plus)) {
      comparisons <- append(comparisons, list(c("0-30", "30+")))
      annotations <- c(annotations, paste0(
        protein_stats$sig_0_30_vs_30plus, "\n(", protein_stats$test_type_0_30_30plus, ")"
      ))
    }
    
    # 
    ## Create plot with lines between same subject:
    p1 <- ggplot(plot_data, aes(x = time_since_vaccination_group, y = npx, group = idnr)) +
      geom_beeswarm(aes( my_col = time_since_vaccination_group, my_x = time_since_vaccination_group))
    pb <- ggplot_build(p1)
    
    p <- ggplot(pb$data[[1]], aes(x = x, y = y, group = group)) +
      geom_line(aes(group = group), col = "grey80", alpha = 0.2) +
      geom_point( aes( color = my_col), size = 3, alpha = 0.6) +
      theme_classic() +
      scale_x_discrete() +
      labs(title = protein, y = "ProtPQN NPX [AU]", x = "Days since closest vaccination", subtitle = "Excluding samples in serocluster 2 and 3 (infected)") +
      geom_boxplot(aes(x = my_x, y = y, group = my_x), outlier.shape = NA, fill = NA) +
      scale_color_manual( values = c("grey", "red", "orange")) +
      guides(color="none")
    
    
    
    # Add significance brackets if we have comparisons
    if(length(comparisons) > 0) {
      p <- p + geom_signif(
        comparisons = comparisons,
        annotations = annotations,
        step_increase = 0.1,
        textsize = 2.5
      )
    }
    
    return(p)
  })
  
  names(plot_list) <- proteins_of_interest
  return(plot_list)
}


time_adjusted_data_si_vaccination <- time_adjusted_data %>%
  left_join(
    olink_bamse_protpqn_blank_rm_intnorm_si %>%
      select(
        samplename, idnr, sampling_phase, sminet_case, sminet_date,
        sampling_time, date1, date2, date3
      ),
    by = "samplename"
  ) %>%
  mutate(
    sampling_time = as.Date(sampling_time),
    dose1_date = as.Date(date1),
    dose2_date = as.Date(date2),
    dose3_date = as.Date(date3)
  )

# select pre- and post-vaccine samples per subject
samples_pre_post <- time_adjusted_data_si_vaccination %>%
  rowwise() %>%
  mutate(
    # Collect dose dates (remove NA)
    dose_dates = list(na.omit(c(dose1_date, dose2_date, dose3_date))),
    # Compute min/max safely
    min_dose_date = if (length(dose_dates) > 0) min(dose_dates) else as.Date(NA),
    max_dose_date = if (length(dose_dates) > 0) max(dose_dates) else as.Date(NA),
    # Find the dose date closest in time (absolute difference)
    closest_dose_date = if (length(dose_dates) > 0) {
      dose_dates[which.min(abs(as.numeric(sampling_time - dose_dates)))]
    } else {
      as.Date(NA)
    },
    # Compute time difference (can be negative if before vaccination)
    time_since_closest_date = as.numeric(sampling_time - closest_dose_date)
  ) %>%
  ungroup() %>%
  group_by(idnr) %>%
  mutate(
    # Classify samples as pre- or post-vaccination
    type = case_when(
      !is.na(min_dose_date) & sampling_time < min_dose_date ~ "pre",
      !is.na(min_dose_date) & sampling_time >= min_dose_date ~ "post",
      TRUE ~ NA_character_
    )
  ) %>%
  # Keep latest pre and earliest post per person
  group_by(idnr, type) %>%
  filter(
    (type == "pre" & sampling_time == max(sampling_time)) |
      (type == "post" & sampling_time == min(sampling_time))
  ) %>%
  ungroup()

# check that we have max two per idnr
samples_pre_post %>%
  count(idnr) %>%
  count(n)



#proteins_of_interest_tad <- proteins_of_interest[-grep("PKLR|GLOD4|DNPH1", proteins_of_interest )]


statistical_results_timeadjust_vaccine <- calculate_protein_stats_vaccine(proteins_of_interest_tad, samples_pre_post)
enhanced_plots_timeadjust_vaccine <- create_plots_with_ggsignif_vaccine(proteins_of_interest_tad, samples_pre_post, statistical_results_timeadjust_vaccine)
```

## Histogram plots with time since infection and time since vaccinaiton (Figure 5 b and c):

```{r}
infected_subj <- time_adjusted_data_si_zscore |>
  filter(idnr %in% sminet_infected_subjects) %>%
  mutate(
    time_since_sminet = as.Date(sampling_time) - as.Date(sminet_date),
    time_since_sminet_group = case_when(
      time_since_sminet < 0 ~ "pre",
      time_since_sminet >= 0 & time_since_sminet <= 30 ~ "0-30",
      time_since_sminet > 30 ~ "30+"
    ),
    time_since_sminet_group = factor(time_since_sminet_group, levels = c("pre", "0-30",  "30+"))
  ) %>%
  filter(Time_Sample_Dose1 < 0) %>%
  filter( ! sampling_phase == "Phase 3")

infected_subj <- infected_subj |> 
  mutate(type = case_when(time_since_sminet < 0 ~ "pre",                                                        time_since_sminet > 0 ~ "post")) |>
  group_by(idnr, type) %>%
  filter(
    (type == "pre" & sampling_time == max(sampling_time)) |
      (type == "post" & sampling_time == min(sampling_time))
  ) %>%
  ungroup() |>
  # Only keep those idnr where we have one pre and one post sample:
  group_by(idnr) %>%
  filter(all(c("pre", "post") %in% type)) %>%
  ungroup()

## Take out vaccinated:
time_adjusted_data_si_zscore_vaccination <- time_adjusted_data_si_zscore %>%
  left_join(
    olink_bamse_protpqn_blank_rm_intnorm_si %>%
      select(
        samplename,
        date1, date2, date3
      ),
    by = "samplename"
  ) %>%
  mutate(
    sampling_time = as.Date(sampling_time),
    dose1_date = as.Date(date1),
    dose2_date = as.Date(date2),
    dose3_date = as.Date(date3)
  )

which.min.pos <- function(x) {
  if (all(x < 0)) return(which.max(x))
  
  min_value <- min(x[x >= 0])
  which(x == min_value)
}

# select pre- and post-vaccine samples per subject
samples_pre_post <- time_adjusted_data_si_zscore_vaccination %>%
  rowwise() %>%
  mutate(
    # Collect dose dates (remove NA)
    dose_dates = list(na.omit(c(dose1_date, dose2_date, dose3_date))),
    # Compute min/max safely
    min_dose_date = if (length(dose_dates) > 0) min(dose_dates) else as.Date(NA),
    max_dose_date = if (length(dose_dates) > 0) max(dose_dates) else as.Date(NA),
    # Find the dose date closest in time (absolute difference)
    closest_dose_date = if (length(dose_dates) > 0) {
      dose_dates[which.min.pos(as.numeric(sampling_time - dose_dates))]
    } else {
      as.Date(NA)
    },
    # Compute time difference (can be negative if before vaccination)
    time_since_closest_date = as.numeric(sampling_time - closest_dose_date)
  ) %>%
  ungroup() %>%
  group_by(idnr) %>%
  mutate(
    # Classify samples as pre- or post-vaccination
    type = case_when(
      !is.na(min_dose_date) & sampling_time < min_dose_date ~ "pre",
      !is.na(min_dose_date) & sampling_time >= min_dose_date ~ "post",
      TRUE ~ NA_character_
    )
  ) %>%
  # Keep latest pre and earliest post per person
  group_by(idnr, type) %>%
  filter(
    (type == "pre" & sampling_time == max(sampling_time)) |
      (type == "post" & sampling_time == min(sampling_time))
  ) %>%
  ungroup() |>
  # Only keep those idnr where we have one pre and one post sample:
  group_by(idnr) %>%
  filter(all(c("pre", "post") %in% type)) %>%
  ungroup()


# Prepare plot data
vaccinated_samples <- samples_pre_post %>% 
  filter(!cluster_maximum_6prot_omi %in% c("2", "3")) |>
  # Make sure we only keep those idnr where we have one pre and one post sample:
  group_by(idnr) %>%
  filter(all(c("pre", "post") %in% type)) %>%
  ungroup()


### Merge and data frames:
infected_subj_merge <- infected_subj |> select(samplename, time_since_sminet, type) |> mutate( time_since_event = time_since_sminet, event = "Infection") |> select( !time_since_sminet)

vaccinated_samples_merge <- vaccinated_samples |> select(samplename, time_since_closest_date, type) |> mutate( time_since_event = time_since_closest_date, event = "Closest vaccination") |> select( !time_since_closest_date)

event_merge <- rbind(infected_subj_merge, vaccinated_samples_merge)


## Plot the histogram
event_means <- event_merge %>%
  group_by(event) %>%
  summarise(mean_time = mean(time_since_event, na.rm = TRUE))


plot1 <- ggplot(event_merge, aes(x = time_since_event, fill = event)) +
  geom_histogram(alpha = 0.6) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1.4) +
  theme_classic() +
  labs(x = "Time since event") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  theme(
    axis.title.x = element_text(size = 16),
    axis.text.x = element_text(size = 14),
    axis.title.y = element_text(size = 16),
    axis.text.y = element_text(size = 14),
    legend.text = element_text(size=14))

event_merge_post_only <- event_merge |> filter( time_since_event >= 0)

event_means <- event_merge_post_only %>%
  group_by(event) %>%
  summarise(mean_time = mean(time_since_event, na.rm = TRUE))

plot2 <- ggplot(event_merge_post_only, aes(x = time_since_event, fill = event)) +
  geom_histogram(alpha = 0.6) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1.4) +
  theme_classic() +
  labs(x = "Time since event") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  theme(
    axis.title.x = element_text(size = 16),
    axis.text.x = element_text(size = 14),
    axis.title.y = element_text(size = 16),
    axis.text.y = element_text(size = 14),
    legend.text = element_text(size=14))

```

### Protein levels in Z-score over time (Figure 5d-g):

```{r}

### Z scores:
protein_since_infection_zscore <- function(proteins_of_interest, data) {
  
  plot_list <- lapply(proteins_of_interest, function(protein) {
    
    # Prepare plot data
    plot_data <- data %>% 
      filter(idnr %in% sminet_infected_subjects) %>%
      mutate(
        time_since_sminet = as.Date(sampling_time) - as.Date(sminet_date),
        time_since_sminet_group = case_when(
          time_since_sminet < 0 ~ "pre",
          time_since_sminet >= 0 & time_since_sminet <= 30 ~ "0-30",
          time_since_sminet > 30 ~ "30+"
        ),
        time_since_sminet_group = factor(time_since_sminet_group, levels = c("pre", "0-30",  "30+"))
      ) %>%
      filter(Time_Sample_Dose1 < 0) %>%
      #  filter( ! sampling_phase == "Phase 3") |>
      pivot_longer(cols = all_of(proteins_of_interest), names_to = "assay", values_to = "z_score") %>%
      filter(assay == protein)
    
    plot_data <- plot_data |> 
      mutate(type = case_when(time_since_sminet < 0 ~ "pre",                                                        time_since_sminet > 0 ~ "post")) |>
      group_by(idnr, type) %>%
      filter(
        (type == "pre" & sampling_time == max(sampling_time)) |
          (type == "post" & sampling_time == min(sampling_time))
      ) %>%
      ungroup() |>
      # Make sure we only keep those idnr where we have one pre and one post sample:
      group_by(idnr) %>%
      filter(all(c("pre", "post") %in% type)) %>%
      ungroup()
    
    
    ymax <- max(abs(c(min(plot_data$z_score, na.rm=TRUE),
                      max(plot_data$z_score, na.rm=TRUE))))
    y_limits <- c(-ymax, ymax)
    
    p <- ggplot(plot_data, aes(x = as.numeric(time_since_sminet), y = z_score, fill = z_score)) +
      geom_point(color = "black", shape = 21, size = 3 ) +
      #scale_fill_viridis(breaks = scales::pretty_breaks(n = 8)) +
      scale_fill_gradient2(
        low = "#09287d",
        mid = "white",
        high = "#7d0916",
        midpoint = 0,
        name = "Z-score"
      ) +
      #      scale_fill_steps2(high = "red", low = "blue", limits = c(-5, 5), midpoint = 0, breaks = c(-3, -2, 2, 3), labels = c("< -3", "-2", "2", ">3")) +
      geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1) +
      geom_hline( yintercept = 0, linetype = "dotdash", color = "black", linewidth = 1) +
      geom_hline( yintercept = 2, linetype = "dotted", color = "black", linewidth = 1) +
      geom_hline( yintercept = -2, linetype = "dotted", color = "black", linewidth = 1) +
      theme_classic() +
      labs(
        title = protein,
        subtitle = "Subjects with PCR+ confirmed infection, no vaccination",
        y = "Z score",
        x = "Days since PCR+"
      ) +
      guides(color = "none") +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 10),
                         limits = y_limits) +
      theme(
        axis.title.x = element_text(size = 16),
        axis.text.x = element_text(size = 14),
        axis.title.y = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        legend.text = element_text(size=14))
    
    
    p1 <- ggplot(plot_data, aes( x = as.factor(1), y = z_score)) +
      #geom_boxplot(color = "black", size = 1, outlier.shape = NA) +
      geom_quasirandom( aes(col = z_score), size = 2) +
      geom_boxplot(color = "black", size = 1, outlier.shape = NA, alpha = 0, width=0.2) +
      theme(axis.title.x=element_blank(),
            axis.text.x=element_blank(),
            axis.ticks.x=element_blank()) +
      scale_color_gradient2(
        low = "#09287d",
        mid = "white",
        high = "#7d0916",
        midpoint = 0,
        name = "Z-score"
      )  +
      theme_classic() +
      #scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 10),
                         limits = y_limits) +
      theme(
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        legend.text = element_text(size=14)
      )
    
    plot_comb <- p1 + p + plot_layout(guides = "collect", design = "
ABB")
    
    return(plot_comb)
    
  })
  names(plot_list) <- proteins_of_interest
  return(plot_list)
  
}


protein_since_infection_all_zscore <- function(proteins_of_interest, data) {
  
  plot_list <- lapply(proteins_of_interest, function(protein) {
    
    # Prepare plot data
    plot_data <- data %>% 
      filter(idnr %in% sminet_infected_subjects) %>%
      mutate(
        time_since_sminet = as.Date(sampling_time) - as.Date(sminet_date)
      ) %>%
      pivot_longer(cols = all_of(proteins_of_interest), names_to = "assay", values_to = "z_score") %>%
      filter(assay == protein)
    
    ymax <- max(abs(c(min(plot_data$z_score, na.rm=TRUE),
                      max(plot_data$z_score, na.rm=TRUE))))
    y_limits <- c(-ymax, ymax)
    
    p <- ggplot(plot_data, aes(x = as.numeric(time_since_sminet), y = z_score, fill = z_score)) +
      geom_point(color = "black", shape = 21, size = 3 ) +
      scale_fill_gradient2(
        low = "#09287d",
        mid = "white",
        high = "#7d0916",
        midpoint = 0,
        name = "Z-score"
      ) +
      #      scale_fill_steps2(high = "red", low = "blue", limits = c(-5, 5), midpoint = 0, breaks = c(-3, -2, 2, 3), labels = c("< -3", "-2", "2", ">3")) +
      geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1) +
      geom_hline( yintercept = 0, linetype = "dotdash", color = "black", linewidth = 1) +
      geom_hline( yintercept = 2, linetype = "dotted", color = "black", linewidth = 1) +
      geom_hline( yintercept = -2, linetype = "dotted", color = "black", linewidth = 1) +
      theme_classic() +
      labs(
        title = protein,
        subtitle = "Subjects with PCR+ confirmed infection and possibly vaccinated subjects",
        y = "Z scores",
        x = "Days since PCR+"
      ) +
      guides(color = "none") +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 10),
                         limits = y_limits)
    
    
    
    p1 <- ggplot(plot_data, aes( x = as.factor(1), y = z_score)) +
      #geom_boxplot(color = "black", size = 1, outlier.shape = NA) +
      geom_quasirandom( aes(col = z_score), size = 2) +
      geom_boxplot(color = "black", size = 1, outlier.shape = NA, alpha = 0, width=0.2) +
      theme(axis.title.x=element_blank(),
            axis.text.x=element_blank(),
            axis.ticks.x=element_blank()) +
      scale_color_gradient2(
        low = "#09287d",
        mid = "white",
        high = "#7d0916",
        midpoint = 0,
        name = "Z-score"
      )  +
      theme_classic() +
      #scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 10),
                         limits = y_limits) +
      theme(
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size=14)
      )
    
    plot_comb <- p1 + p + plot_layout(guides = "collect", design = "
ABB")
    
    return(plot_comb)
    
  })
  names(plot_list) <- proteins_of_interest
  return(plot_list)
  
}

## z-score transformation:
time_adjusted_data_si_zscore <- time_adjusted_data_si %>%
  mutate(
    across(
      all_of(proteins_of_interest_tad),
      ~ ( . - mean(., na.rm = TRUE) ) / sd(., na.rm = TRUE)
    )
  )

plots_zscores_I <- protein_since_infection_zscore(proteins_of_interest_tad, time_adjusted_data_si_zscore)

plots_zscores_VI <- protein_since_infection_all_zscore(proteins_of_interest_tad, time_adjusted_data_si_zscore)


#### Vaccinated people only

time_adjusted_data_si_zscore_vaccination <- time_adjusted_data_si_zscore %>%
  left_join(
    olink_bamse_protpqn_blank_rm_intnorm_si %>%
      select(
        samplename,
        date1, date2, date3
      ),
    by = "samplename"
  ) %>%
  mutate(
    sampling_time = as.Date(sampling_time),
    dose1_date = as.Date(date1),
    dose2_date = as.Date(date2),
    dose3_date = as.Date(date3)
  )

which.min.pos <- function(x) {
  if (all(x < 0)) return(which.max(x))
  
  min_value <- min(x[x >= 0])
  which(x == min_value)
}

# select pre- and post-vaccine samples per subject
samples_pre_post <- time_adjusted_data_si_zscore_vaccination %>%
  rowwise() %>%
  mutate(
    # Collect dose dates (remove NA)
    dose_dates = list(na.omit(c(dose1_date, dose2_date, dose3_date))),
    # Compute min/max safely
    min_dose_date = if (length(dose_dates) > 0) min(dose_dates) else as.Date(NA),
    max_dose_date = if (length(dose_dates) > 0) max(dose_dates) else as.Date(NA),
    # Find the dose date closest in time (absolute difference)
    closest_dose_date = if (length(dose_dates) > 0) {
      dose_dates[which.min.pos(as.numeric(sampling_time - dose_dates))]
    } else {
      as.Date(NA)
    },
    # Compute time difference (can be negative if before vaccination)
    time_since_closest_date = as.numeric(sampling_time - closest_dose_date)
  ) %>%
  ungroup() %>%
  group_by(idnr) %>%
  mutate(
    # Classify samples as pre- or post-vaccination
    type = case_when(
      !is.na(min_dose_date) & sampling_time < min_dose_date ~ "pre",
      !is.na(min_dose_date) & sampling_time >= min_dose_date ~ "post",
      TRUE ~ NA_character_
    )
  ) %>%
  # Keep latest pre and earliest post per person
  group_by(idnr, type) %>%
  filter(
    (type == "pre" & sampling_time == max(sampling_time)) |
      (type == "post" & sampling_time == min(sampling_time))
  ) %>%
  ungroup() |>
  # Make sure we only keep those idnr where we have one pre and one post sample:
  group_by(idnr) %>%
  filter(all(c("pre", "post") %in% type)) %>%
  ungroup()

# check that we have max two per idnr
samples_pre_post %>%
  count(idnr) %>%
  count(n)

protein_since_vaccination_zcore <- function(proteins_of_interest, data) {
  
  plot_list <- lapply(proteins_of_interest, function(protein) {
    
    # Prepare plot data
    plot_data <- data %>% 
      filter(!cluster_maximum_6prot_omi %in% c("2", "3")) |>
      pivot_longer(cols = all_of(proteins_of_interest), names_to = "assay", values_to = "z_score") %>%
      # group_by(assay) %>%
      #    mutate(
      #      z_score = (npx - mean(npx, na.rm = TRUE)) / sd(npx, na.rm = TRUE)
      #    ) %>%
      filter(assay == protein) |>
      # Make sure we only keep those idnr where we have one pre and one post sample:
      group_by(idnr) %>%
      filter(all(c("pre", "post") %in% type)) %>%
      ungroup()
    
    # # Prepare significance 
    ymax <- max(abs(c(min(plot_data$z_score, na.rm=TRUE),
                      max(plot_data$z_score, na.rm=TRUE))))
    y_limits <- c(-ymax, ymax)
    
    
    p <- ggplot(plot_data, aes(x = as.numeric(time_since_closest_date), y = z_score, fill = z_score)) +
      geom_point(color = "black", shape = 21, size = 3 ) +
      scale_fill_gradient2(
        low = "#09287d",
        mid = "white",
        high = "#7d0916",
        midpoint = 0,
        name = "Z-score"
      )  +
      #scale_fill_steps2(high = "red", low = "blue", limits = c(-5, 5), midpoint = 0, breaks = c(-3, -2, 2, 3), labels = c("< -3", "-2", "2", ">3")) +
      #                 scale_fill_gradient2(
      #   low = "#4575B4",
      #   mid = "white",
      #   high = "#D73027",
      #   midpoint = 0,
      #   name = "Z-score"
      # )+
      #scale_fill_viridis(breaks = scales::pretty_breaks(n = 8)) +
      geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1) +
      geom_hline( yintercept = 0, linetype = "dotdash", color = "black", linewidth = 1) +
      geom_hline( yintercept = 2, linetype = "dotted", color = "black", linewidth = 1) +
      geom_hline( yintercept = -2, linetype = "dotted", color = "black", linewidth = 1) +
      theme_classic() +
      labs(
        title = protein,
        subtitle = "Excluding samples in serocluster 2 and 3 (infected)",
        y = "Z-score",
        x = "Days since closest vaccination"
      ) +
      # guides(color = "none") +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 10),
                         limits = y_limits) +
      theme(
        axis.title.x = element_text(size = 16),
        axis.text.x = element_text(size = 14),
        axis.title.y = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        legend.text = element_text(size=14))
    
    
    p1 <- ggplot(plot_data, aes( x = as.factor(1), y = z_score)) +
      #geom_boxplot(color = "black", size = 1, outlier.shape = NA) +
      geom_quasirandom( aes(col = z_score), size = 2) +
      geom_boxplot(color = "black", size = 1, outlier.shape = NA, alpha = 0, width=0.2) +
      theme(axis.title.x=element_blank(),
            axis.text.x=element_blank(),
            axis.ticks.x=element_blank()) +
      scale_color_gradient2(
        low = "#09287d",
        mid = "white",
        high = "#7d0916",
        midpoint = 0,
        name = "Z-score"
      )  +
      theme_classic() +
      #scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 10),
                         limits = y_limits) +
      theme(
        axis.title.x = element_blank(),
        #axis.text.x = element_text(size = 14),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size=14)
      )
    
    plot_comb <- p1 + p + plot_layout(guides = "collect", design = "
ABB")
    
    return(plot_comb)
    
    
    
  })
  names(plot_list) <- proteins_of_interest
  return(plot_list)
  
}

plots_zscores_V <- protein_since_vaccination_zcore(proteins_of_interest_tad, samples_pre_post)





plot_list <- lapply( 1:length(proteins_of_interest_tad), function( n){
  
  protein <- proteins_of_interest_tad[n]
  
  # Combine the z-score values across both datasets
  combined_z_limits <- range(
    time_adjusted_data_si_zscore[protein],
    na.rm = TRUE
  )
  
  # Optionally, extend slightly for nicer color distribution
  combined_z_limits[1] <- floor(combined_z_limits[1])
  combined_z_limits[2] <- ceiling(combined_z_limits[2])
  
  
  # Update the fill aesthetic mapping to use z_bin instead of z_score
  plots_zscores_I[[n]]$mapping$fill <- aes(fill = z_score)$fill
  plots_zscores_V[[n]]$mapping$fill <- aes(fill = z_score)$fill
  
  p_I <- plots_zscores_I[[n]]
  p_V <- plots_zscores_V[[n]]
  
  
  (wrap_plots(p_I) + wrap_plots(p_V)) +
    plot_layout(guides = "collect") &
    scale_y_continuous(limits = combined_z_limits)
  
})

```

### Comparing proteins in relation to infection and vaccination (Fig. S19):

```{r}

# Comparing protein levels in previously infected versus seronegative subjects (Fig. S19a): 
#  COMPARE 11 and 22 at baseline:
n_runs <- 100
fdr_cutoff <- 0.05
seed_base <- 123

library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)
library(progress)

# Precompute group sizes
n12 <- time_adjusted_data_for_lm %>%
  filter(sampling_phase != "Phase 3", clust_traj12 == "22") %>%
  distinct(idnr) %>%
  nrow()

# Get random IDs in group 11
group11_ids <- time_adjusted_data_for_lm %>%
  filter(sampling_phase != "Phase 3", clust_traj12 == "11") %>%
  distinct(idnr) %>%
  pull(idnr)

# Function for bootstrap iteration
run_one_bootstrap_within_phase <- function(i) {
  set.seed(seed_base + i)
  
  # randomly sample n12 subjects from group11 (WITH replacement for bootstrap)
  sampled_11 <- sample(group11_ids, size = n12, replace = TRUE)
  
  # build data_list for analysis: keep group12 and the sampled group11 subjects
  data_subset <- time_adjusted_data_for_lm %>%
    filter(sampling_phase != "Phase 3") %>%
    filter(clust_traj12 == "22" | idnr %in% sampled_11)
  
  data_list <- list(
    values = data_subset %>%
      dplyr::select(sample_id, all_of(olink_proteins_ta)),     # adjust variable name if needed
    sinfo = data_subset %>%
      dplyr::select(idnr, sample_id, sampling_phase, clust_traj12, male)
  )
  
  # Run the delta-delta analysis (difference in change between Phase1 and Phase2 for each group)
  lm_dd <- lm_analysis(
    data_list,
    "sample_id",
    predictor = "clust_traj12",
    covar = "male",
    stratify_by = "sampling_phase"
    # difference_col = "sampling_phase",
    #  individual_id = "idnr",
    # differences_between = c("Phase 1", "Phase 2")
  )
  
  # lm_dd$results expected to be a data.frame with columns: assay, estimate, p_value, fdr
  res <- lm_dd$results %>%
    mutate(iter = i) %>%
    dplyr::select(assay, sampling_phase, median_diff, estimate, p_value, fdr, iter)
  
  # Return list of significant assays this run
  sig_assays_p1 <- res %>%filter(sampling_phase == "Phase 1") %>% filter(fdr < fdr_cutoff) %>% pull(assay)
  
  sig_assays_p2 <- res %>%filter(sampling_phase == "Phase 2") %>% filter(fdr < fdr_cutoff) %>% pull(assay)
  
  list(results = res, significant_p1 = sig_assays_p1, significant_p2 = sig_assays_p2)
}

# Run many iterations (serial)
pb <- progress_bar$new(total = n_runs)
all_iter_results <- vector("list", n_runs)
for (i in seq_len(n_runs)) {
  iter_out <- tryCatch({
    out <- run_one_bootstrap_within_phase(i)
    pb$tick()
    out
  }, error = function(e) {
    pb$tick()
    message("Error in iteration ", i, ": ", e$message)
    list(results = tibble(assay = character(), median_diff = numeric(), estimate = numeric(), p_value = numeric(), fdr = numeric(), iter = integer()),
         significant_p1 = character(0),
         significant_p2 = character(0))
  })
  all_iter_results[[i]] <- iter_out
}

# Aggregate outputs
# Frequency table: how many times each assay was significant
all_sig_lists_p1 <- map(all_iter_results, "significant_p1")
all_sig_flat_p1 <- unlist(all_sig_lists_p1)
protein_counts <- sort(table(all_sig_flat_p1), decreasing = TRUE)
protein_counts_p1_df <- tibble(
  assay = names(protein_counts),
  count = as.integer(protein_counts),
  freq = count / n_runs
)


all_sig_lists_p2 <- map(all_iter_results, "significant_p2")
all_sig_flat_p2 <- unlist(all_sig_lists_p2)
protein_counts <- sort(table(all_sig_flat_p2), decreasing = TRUE)
protein_counts_p2_df <- tibble(
  assay = names(protein_counts),
  count = as.integer(protein_counts),
  freq = count / n_runs
)

# Collect all per-iteration results (estimates, p-values)
all_results_df <- all_iter_results %>%
  map("results") %>%
  list_rbind()

# Compute summary statistics and 95% confidence intervals per assay
summary_df <- all_results_df %>%
  group_by(sampling_phase, assay) %>%
  summarise(
    median_median_diff = median( median_diff, na.rm = TRUE),
    median_estimate = median(estimate, na.rm = TRUE),
    mean_estimate = mean(estimate, na.rm = TRUE),
    lower_95 = quantile(estimate, 0.025, na.rm = TRUE),
    upper_95 = quantile(estimate, 0.975, na.rm = TRUE),
    prop_pos = mean(estimate > 0, na.rm = TRUE),
    prop_neg = mean(estimate < 0, na.rm = TRUE),
    prop_sig = mean(fdr < fdr_cutoff, na.rm = TRUE),
    n_non_na = sum(!is.na(estimate)),
    fdr_median = median(fdr),
    fdr_mean = mean(fdr),
    .groups = "drop"
  ) %>%
  arrange(desc(prop_sig), desc(prop_pos))

summary_df_1122 <- summary_df

### Plot Volcano plot:
volcano_1122_baseline <- ggplot(summary_df_1122, aes( x = median_median_diff, y = -log10(fdr_median))) +
  geom_point( aes( color = prop_sig, size = prop_sig)) +
  facet_wrap(~sampling_phase) +
  theme_classic() +
  #theme_bw() +
  geom_text_repel(
    data = subset(summary_df_1122, prop_sig > 0.1),
    aes(label = str_remove(assay, "_Phase \\d")),  # Use the `protein` column as the label
    color = "black",
    box.padding = 0.5,
    max.overlaps = Inf
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  #colr_scale +
  scale_x_continuous(limits = center_limits()) +
  scale_color_gradientn(
    colors = c("grey80", wes_palette("Zissou1", 100, type = "continuous")),
    values = scales::rescale(c(0, 0.0001, 1)),  # Grey for 0, then gradient starts
    limits = c(0, 1),
    name = "Proportion\nsignificant"
  ) +
  scale_y_continuous(breaks = round(seq(0, max(-log10(summary_df$fdr_median)+2), by = 1),1)) +
  labs(
    title = "Subjects who were infected in phase 2 vs naive",
    subtitle = "Bootstrapping naive group for 100 iterations",
    x = "Median difference between groups",
    y = "-log10(FDR)"
  ) +
  geom_vline( xintercept = 0, linetype = "dashed", color = "grey") +
  theme(plot.title = element_text(hjust = 0.5))



#---------------------------------------------------------------------------------------------------------
# Comparing protein levels in previously infected versus newly infected subjects (Fig. S19b):  

# 1 -> 2 vs 2 -> 2
data_list_12_22 <- list(
  values = time_adjusted_data_for_lm %>%
    filter(sampling_phase != "Phase 3") %>%
    filter(clust_traj12 %in% c("12", "22")) %>%
    dplyr::select(sample_id, !!olink_proteins_ta),
  
  sinfo = time_adjusted_data_for_lm %>%
    filter(sampling_phase != "Phase 3") %>%
    filter(clust_traj12 %in% c("12", "22")) %>%
    dplyr::select(idnr, sample_id, sampling_phase, clust_traj12, male)
)

# lm between 12 and 22 at each sampling phase
lm_12_22 <- lm_analysis(data_list_12_22, "sample_id", predictor = "clust_traj12", covar = "male",
                        stratify_by = "sampling_phase")

plts_12_22 <- plot_lm_results(lm_12_22, sep_by_colr = F)

plts_12_22$volcano


#---------------------------------------------------------------------------------------------------------
# Comparing protein levels in previously infected who get vaccinated versus newly infected AND vaccinated subjects (Fig. S19c): 

# between phase 2 and 3, cluster 1 -> 3 and 2 -> 3
data_list_13_23 <- list(
  values = time_adjusted_data_for_lm %>%
    filter(sampling_phase != "Phase 1") %>%
    filter(clust_traj23 %in% c("13", "23")) %>%
    dplyr::select(sample_id, !!olink_proteins_ta),
  
  sinfo = time_adjusted_data_for_lm %>%
    filter(sampling_phase != "Phase 1") %>%
    filter(clust_traj23 %in% c("13", "23")) %>%
    dplyr::select(idnr, sample_id, sampling_phase, clust_traj23, male)
)

# lm between 13 and 23 at each sampling phase
lm_13_23 <- lm_analysis(data_list_13_23, "sample_id", predictor = "clust_traj23", covar = "male",
                        stratify_by = "sampling_phase")

plts_13_23 <- plot_lm_results(lm_13_23, sep_by_colr = F)

plts_13_23$volcano


#---------------------------------------------------------------------------------------------------------
# Previously infected subjects with waning anti-N antibodies who became vaccinated prior to Phase 3 (Fig. S19d):
# 14 and 24 groups at baseline each phase

n_runs <- 100
fdr_cutoff <- 0.05
seed_base <- 123

library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)
library(progress)

# Precompute group sizes
n24 <- time_adjusted_data_for_lm %>%
  filter(sampling_phase != "Phase 1", clust_traj23 == "24") %>%
  distinct(idnr) %>%
  nrow()

# Get random IDs in group 11
group14_ids <- time_adjusted_data_for_lm %>%
  filter(sampling_phase != "Phase 1", clust_traj23 == "14") %>%
  distinct(idnr) %>%
  pull(idnr)


# Function for bootstrap iteration
run_one_bootstrap_within_phase <- function(i) {
  set.seed(seed_base + i)
  
  # randomly sample n12 subjects from group11 (WITH replacement for bootstrap)
  sampled_11 <- sample(group14_ids, size = n24, replace = TRUE)
  
  # build data_list for analysis: keep group12 and the sampled group11 subjects
  data_subset <- time_adjusted_data_for_lm %>%
    filter(sampling_phase != "Phase 1") %>%
    filter(clust_traj23 == "24" | idnr %in% sampled_11)
  
  data_list <- list(
    values = data_subset %>%
      dplyr::select(sample_id, all_of(olink_proteins_ta)),     # adjust variable name if needed
    sinfo = data_subset %>%
      dplyr::select(idnr, sample_id, sampling_phase, clust_traj23, male)
  )
  
  # Run the delta-delta analysis (difference in change between Phase1 and Phase2 for each group)
  lm_dd <- lm_analysis(
    data_list,
    "sample_id",
    predictor = "clust_traj23",
    covar = "male",
    stratify_by = "sampling_phase"
  )
  
  # lm_dd$results expected to be a data.frame with columns: assay, estimate, p_value, fdr
  res <- lm_dd$results %>%
    mutate(iter = i) %>%
    dplyr::select(assay, sampling_phase, median_diff, estimate, p_value, fdr, iter)
  
  # Return list of significant assays this run
  sig_assays_p2 <- res %>%filter(sampling_phase == "Phase 2") %>% filter(fdr < fdr_cutoff) %>% pull(assay)
  
  sig_assays_p3 <- res %>%filter(sampling_phase == "Phase 3") %>% filter(fdr < fdr_cutoff) %>% pull(assay)
  
  list(results = res, significant_p2 = sig_assays_p2, significant_p3 = sig_assays_p3)
}

# Run many iterations (serial)
pb <- progress_bar$new(total = n_runs)
all_iter_results <- vector("list", n_runs)
for (i in seq_len(n_runs)) {
  iter_out <- tryCatch({
    out <- run_one_bootstrap_within_phase(i)
    pb$tick()
    out
  }, error = function(e) {
    pb$tick()
    message("Error in iteration ", i, ": ", e$message)
    list(results = tibble(assay = character(), median_diff = numeric(), estimate = numeric(), p_value = numeric(), fdr = numeric(), iter = integer()),
         significant_p2 = character(0),
         significant_p3 = character(0))
  })
  all_iter_results[[i]] <- iter_out
}

# Aggregate outputs
# Frequency table: how many times each assay was significant
all_sig_lists_p2 <- map(all_iter_results, "significant_p2")
all_sig_flat_p2 <- unlist(all_sig_lists_p2)
protein_counts <- sort(table(all_sig_flat_p2), decreasing = TRUE)
protein_counts_p2_df <- tibble(
  assay = names(protein_counts),
  count = as.integer(protein_counts),
  freq = count / n_runs
)


all_sig_lists_p3 <- map(all_iter_results, "significant_p3")
all_sig_flat_p3 <- unlist(all_sig_lists_p3)
protein_counts <- sort(table(all_sig_flat_p3), decreasing = TRUE)
protein_counts_p3_df <- tibble(
  assay = names(protein_counts),
  count = as.integer(protein_counts),
  freq = count / n_runs
)



# Collect all per-iteration results (estimates, p-values)
all_results_df <- all_iter_results %>%
  map("results") %>%
  list_rbind()

# Compute summary statistics and 95% confidence intervals per assay
summary_df <- all_results_df %>%
  group_by(sampling_phase, assay) %>%
  summarise(
    median_median_diff = median( median_diff, na.rm = TRUE),
    median_estimate = median(estimate, na.rm = TRUE),
    mean_estimate = mean(estimate, na.rm = TRUE),
    lower_95 = quantile(estimate, 0.025, na.rm = TRUE),
    upper_95 = quantile(estimate, 0.975, na.rm = TRUE),
    prop_pos = mean(estimate > 0, na.rm = TRUE),
    prop_neg = mean(estimate < 0, na.rm = TRUE),
    prop_sig = mean(fdr < fdr_cutoff, na.rm = TRUE),
    n_non_na = sum(!is.na(estimate)),
    fdr_median = median(fdr),
    fdr_mean = mean(fdr),
    .groups = "drop"
  ) %>%
  arrange(desc(prop_sig), desc(prop_pos))


summary_df_2414 <- summary_df

### Plot Volcano plot:
volcano_2414_baseline <- ggplot(summary_df_2414, aes( x = median_median_diff, y = -log10(fdr_median))) +
  geom_point( aes( color = prop_sig, size = prop_sig)) +
  facet_wrap(~sampling_phase) +
  theme_classic() +
  #theme_bw() +
  geom_text_repel(
    data = subset(summary_df_2414, prop_sig > 0.1),
    aes(label = str_remove(assay, "_Phase \\d")),  # Use the `protein` column as the label
    color = "black",
    box.padding = 0.5,
    max.overlaps = Inf
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  #colr_scale +
  scale_x_continuous(limits = center_limits()) +
  scale_color_gradientn(
    colors = c("grey80", wes_palette("Zissou1", 100, type = "continuous")),
    values = scales::rescale(c(0, 0.0001, 1)),  # Grey for 0, then gradient starts
    limits = c(0, 1),
    name = "Proportion\nsignificant"
  ) +
  #scale_y_continuous(breaks = round(seq(0, max(-log10(summary_df$fdr_median)+2), by = 1),1)) +
  scale_y_continuous(limits = c(0, 6), breaks = round(seq(0, 6, by = 1),1)) +
  labs(
    title = "Subjects who were previously infected vs naive who got vaccinated",
    subtitle = "Bootstrapping naive group for 100 iterations",
    x = "Median difference between groups",
    y = "-log10(FDR)"
  ) +
  geom_vline( xintercept = 0, linetype = "dashed", color = "grey") +
  theme(plot.title = element_text(hjust = 0.5))


#---------------------------------------------------------------------------------------------------------
# Protein levels between phases for different cluster changing groups (S19e-h):

## 1-> 2 vs 2-2:
lap3_12_vs_22 <- left_join( data_list_12_22$values, data_list_12_22$sinfo, by = "sample_id") |>
  ggplot( aes( y = LAP3, x = sampling_phase)) +
  geom_line(aes(group = idnr), alpha = 0.3, color = "grey80") +
  geom_quasirandom(alpha = 0.7, size = 1.5, color = "grey60") +
  #geom_beeswarm(alpha = 0.5, size = 3, color = "grey70") +
  geom_boxplot( fill = NA, outlier.shape = NA) +
  facet_wrap( ~ clust_traj12) +
  theme_classic(12) +
  stat_compare_means( size = 2.1) +
  labs( y = "NPX [AU]", x = "sampling phase", title = "LAP3") +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(size = 8))


LGALS9_12_vs_22 <- left_join( data_list_12_22$values, data_list_12_22$sinfo, by = "sample_id") |>
  ggplot( aes( y = LGALS9, x = sampling_phase)) +
  geom_line(aes(group = idnr), alpha = 0.3, color = "grey80") +
  geom_quasirandom(alpha = 0.7, size = 1.5, color = "grey60") +
  #geom_beeswarm(alpha = 0.5, size = 3, color = "grey70") +
  geom_boxplot( fill = NA, outlier.shape = NA) +
  facet_wrap( ~ clust_traj12) +
  theme_classic(12) +
  stat_compare_means( size = 2.1) +
  labs( y = "NPX [AU]", x = "sampling phase", title = "LGALS9") +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(size = 8))

traf2_12_vs_22 <- left_join( data_list_12_22$values, data_list_12_22$sinfo, by = "sample_id") |>
  ggplot( aes( y = TRAF2, x = sampling_phase)) +
  #geom_beeswarm(alpha = 0.5, size = 3, color = "grey70") +
  geom_line(aes(group = idnr), alpha = 0.3, color = "grey80") +
  geom_quasirandom(alpha = 0.7, size = 1.5, color = "grey60") +
  geom_boxplot( fill = NA, outlier.shape = NA) +
  facet_wrap( ~ clust_traj12) +
  theme_classic(12) +
  stat_compare_means( size = 2.1) +
  labs( y = "NPX [AU]", x = "sampling phase", title = "TRAF2") +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(size = 8))

## 1-3 vs 2-3:
lap3_13_vs_23 <- left_join( data_list_13_23$values, data_list_13_23$sinfo, by = "sample_id") |>
  ggplot( aes( y = LAP3, x = sampling_phase)) +
  geom_line(aes(group = idnr), alpha = 0.3, color = "grey80") +
  geom_quasirandom(alpha = 0.7, size = 1.5, color = "grey60") +
  #geom_beeswarm(alpha = 0.5, size = 3, color = "grey70") +
  geom_boxplot( fill = NA, outlier.shape = NA) +
  facet_wrap( ~ clust_traj23) +
  theme_classic(12) +
  stat_compare_means( size = 2.1) +
  labs( y = "NPX [AU]", x = "sampling phase", title = "LAP3") +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(size = 8))

```

```{r}
df <- read.csv("sampling and vaccination dates_fig1b.csv")

library(dplyr)
library(tidyr)
library(lubridate)

df_summarized <- df |> 
  # 1. Convert character strings to Date class
  mutate(across(c(sampling_time, date1, date2, date3), ymd)) |> 
  
  # 2. Invalidate doses that occurred AFTER sampling_time
  mutate(
    date1 = if_else(date1 <= sampling_time, date1, NA_Date_),
    date2 = if_else(date2 <= sampling_time, date2, NA_Date_),
    date3 = if_else(date3 <= sampling_time, date3, NA_Date_)
  ) |> 
  
  # 3. Pivot dose dates to long format to extract month/year
  pivot_longer(
    cols = c(date1, date2, date3),
    names_to = "dose",
    values_to = "vax_date"
  ) |> 
  
  # 4. Keep valid vaccination dates and truncate to Year-Month (YYYY-MM-01)
  filter(!is.na(vax_date)) |> 
  mutate(date = floor_date(vax_date, unit = "month")) |> 
  
  # 5. Count samples per month per dose level
  count(date, dose) |> 
  
  # 6. Pivot back wide so each dose has its own column
  pivot_wider(
    names_from = dose,
    values_from = n,
    values_fill = 0
  ) |> 
  
  # 7. Rename columns and ensure date display format (YYYY-MM)
  rename(
    vaccine_dose_1 = date1,
    vaccine_dose_2 = date2,
    vaccine_dose_3 = date3
  ) |> 
  mutate(date = format(date, "%Y-%m")) |> 
  arrange(date)
```
